---
title: "00 — Data Engineering"
subtitle: "PGA Tour Distance & Speed Analytics"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    df_print: paged
---

# Overview

This notebook is the single source of truth for all data cleaning and variable
engineering in this project. It reads the two raw datasets, applies cleaning
decisions, engineers all derived variables, and writes processed outputs to
`data/processed/`. Every subsequent analysis notebook loads from those outputs
rather than re-processing the raw data.

**Inputs:**
- `data/raw/df_complete.csv` — Player-season records, 2004–2025
- `data/raw/trends_acc_dist_1987_to_2025.csv` — Distance and accuracy, 1987–2025

**Outputs:**
- `data/processed/df_main.rds` — Cleaned and engineered df_complete
- `data/processed/df_trends.rds` — Cleaned df_trends

**Note on survival data:** The survival analysis dataset (`df_survival.rds`) is
constructed separately in `00b_survival_prep.Rmd` once we have sufficient
familiarity with career gap patterns in the data to make defensible censoring
decisions.

---

# 1. Setup

```{r setup, message=FALSE, warning=FALSE}
# Core
library(tidyverse)
library(here)

# Utilities
library(broom)

# Source shared config (paths, constants, regulatory events, plot theme)
source(here::here("config.R"))

# Reproducibility
set.seed(2025)

# Global knitr options
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 10,
  fig.height = 6
)
```

---

# 2. Load Raw Data

```{r load-data}
df_raw <- read_csv(
  here(PATH_RAW, "df_complete.csv"),
  show_col_types = FALSE
)

df_trends_raw <- read_csv(
  here(PATH_RAW, "trends_acc_dist_1987_to_2025.csv"),
  show_col_types = FALSE
)
```

## 2.1 Initial inspection — df_complete

We rely on `player_name` + `year` as the unique key for this dataset in every
downstream `left_join`. Confirm that holds before trusting any join result.

```{r key-check-complete}
n_dupes_complete <- df_raw %>%
  count(player_name, year) %>%
  filter(n > 1) %>%
  nrow()

cat("Rows:                    ", nrow(df_raw), "\n")
cat("Distinct player+year keys:", n_distinct(df_raw$player_name, df_raw$year), "\n")
cat("Duplicate player+year keys:", n_dupes_complete, "\n")

stopifnot(n_dupes_complete == 0)
```

```{r inspect-complete}
glimpse(df_raw)
```

```{r inspect-complete-summary}
cat("Rows:           ", nrow(df_raw), "\n")
cat("Unique players: ", n_distinct(df_raw$player_name), "\n")
cat("Year range:     ", min(df_raw$year), "–", max(df_raw$year), "\n")
cat("\nNA counts:\n")
df_raw %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "column", values_to = "n_missing") %>%
  filter(n_missing > 0) %>%
  print()
```

Sanity-check that key measures fall within plausible bounds for PGA Tour
play. This is a coarse guard against data-entry errors, not a claim about
true physical limits — values outside these ranges are reported, not
dropped.

```{r range-check-complete}
df_raw %>%
  summarise(
    accuracy_out_of_range = sum(accuracy < 0 | accuracy > 100, na.rm = TRUE),
    distance_out_of_range = sum(distance < 200 | distance > 350, na.rm = TRUE),
    ball_speed_out_of_range = sum(ball_speed < 140 | ball_speed > 200, na.rm = TRUE),
    chs_out_of_range = sum(chs < 90 | chs > 140, na.rm = TRUE),
    carry_out_of_range = sum(carry_distance < 200 | carry_distance > 320, na.rm = TRUE),
    age_out_of_range = sum(age < 18 | age > 55, na.rm = TRUE)
  ) %>%
  print()

# Age specifically: print the offending rows, since an out-of-range age
# usually points to a birth_year data error rather than a true outlier
# (found via Project 6: "Patrick Sheehan" has 6 rows sharing a birth_year
# that implies ages 12-20, inconsistent with that same row's turned_pro -
# almost certainly two different real people merged under one name).
df_raw %>%
  filter(age < 18 | age > 55) %>%
  select(player_name, year, age, birth_year, turned_pro)
```

## 2.2 Initial inspection — df_trends

```{r key-check-trends}
n_dupes_trends <- df_trends_raw %>%
  count(player, year) %>%
  filter(n > 1) %>%
  nrow()

cat("Rows:                    ", nrow(df_trends_raw), "\n")
cat("Distinct player+year keys:", n_distinct(df_trends_raw$player, df_trends_raw$year), "\n")
cat("Duplicate player+year keys:", n_dupes_trends, "\n")

stopifnot(n_dupes_trends == 0)
```

```{r inspect-trends}
glimpse(df_trends_raw)
```

```{r inspect-trends-summary}
cat("Rows:           ", nrow(df_trends_raw), "\n")
cat("Unique players: ", n_distinct(df_trends_raw$player), "\n")
cat("Year range:     ", min(df_trends_raw$year), "–", max(df_trends_raw$year), "\n")
cat("\nNA counts:\n")
df_trends_raw %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "column", values_to = "n_missing") %>%
  filter(n_missing > 0) %>%
  print()
```

```{r range-check-trends}
df_trends_raw %>%
  summarise(
    accuracy_out_of_range = sum(accuracy < 0 | accuracy > 100, na.rm = TRUE),
    distance_out_of_range = sum(distance < 200 | distance > 350, na.rm = TRUE)
  ) %>%
  print()
```

---

# 3. Clean df_complete

## 3.1 Column names and types

```{r clean-colnames}
df_clean <- df_raw %>%
  # Fix column name with space
  rename(rookie_year = `rookie year`) %>%
  # Ensure player_name is character, year is integer
  mutate(
    player_name = as.character(player_name),
    year        = as.integer(year)
  )

# Confirm rename worked
names(df_clean)
```

## 3.2 Handle the pre-2007 speed data gap

Ball speed, CHS, and carry distance have ~594 NAs concentrated in 2004–2006,
before these metrics were tracked by ShotLink. We do not impute these values.
Instead we document the gap and note the effective start year for speed analyses.

```{r speed-na-check}
df_clean %>%
  group_by(year) %>%
  summarise(
    n             = n(),
    na_ball_speed = sum(is.na(ball_speed)),
    na_chs        = sum(is.na(chs)),
    na_carry      = sum(is.na(carry_distance))
  ) %>%
  filter(na_ball_speed > 0) %>%
  print()
```

The `SPEED_ERA_START` constant in `config.R` is set to `2007`. Any analysis
involving ball speed or CHS should filter to `year >= SPEED_ERA_START`.

## 3.2b Missing rookie_year / turned_pro

`turned_pro` is a weak proxy for PGA Tour entry regardless of missingness:
checked directly against the data, the median gap between turning pro and a
player's first PGA Tour season is 7 years, and 62% of players have a gap
greater than 5 years (grinding through other tours first). Combined with
`rookie_year`'s incompleteness below, neither field is used as the basis for
any engineered variable in this notebook — `years_on_tour` and
`entry_cohort` are instead built from the cross-referenced `entry_year`
(section 3.4). `rookie_year`/`turned_pro` are kept as raw columns for
reference only.

```{r rookie-year-na-check}
missing_rookie_year <- df_clean %>%
  filter(is.na(rookie_year))

cat("Rows missing rookie_year:   ", nrow(missing_rookie_year), "\n")
cat("Players missing rookie_year:", n_distinct(missing_rookie_year$player_name), "\n")
cat("Rows missing turned_pro:    ", sum(is.na(df_clean$turned_pro)), "\n")
```

## 3.3 Minimum seasons filter

Players appearing in fewer than `MIN_SEASONS` (2) seasons are excluded from
longitudinal analyses. We flag rather than drop them here, preserving the
option to include single-season players in purely cross-sectional work.

```{r min-seasons}
seasons_per_player <- df_clean %>%
  count(player_name, name = "n_seasons")

df_clean <- df_clean %>%
  left_join(seasons_per_player, by = "player_name") %>%
  mutate(multi_season = n_seasons >= MIN_SEASONS)

cat("Players with 1 season only: ",
    sum(seasons_per_player$n_seasons < MIN_SEASONS), "\n")
cat("Players with 2+ seasons:    ",
    sum(seasons_per_player$n_seasons >= MIN_SEASONS), "\n")
```

## 3.4 True entry year (cross-referenced with trends data)

`df_complete` only covers the strokes-gained era (2004+), so a player's
first row here isn't necessarily their true PGA Tour debut — some players
were already playing for years beforehand. `trends_acc_dist_1987_to_2025.csv`
carries simpler stats (distance/accuracy only, no strokes-gained) but goes
back to 1987, so we take the earliest year found in *either* source as the
best available estimate of a player's true tour entry.

```{r true-entry-year}
first_year_complete <- df_clean %>%
  group_by(player_name) %>%
  summarise(first_year_complete = min(year), .groups = "drop")

first_year_trends <- df_trends_raw %>%
  rename(player_name = player) %>%
  group_by(player_name) %>%
  summarise(first_year_trends = min(year), .groups = "drop")

TRENDS_START_YEAR <- min(df_trends_raw$year)

entry_year_lookup <- first_year_complete %>%
  left_join(first_year_trends, by = "player_name") %>%
  mutate(
    entry_year     = pmin(first_year_complete, first_year_trends, na.rm = TRUE),
    left_truncated = entry_year == TRENDS_START_YEAR
  ) %>%
  select(player_name, entry_year, left_truncated)

df_clean <- df_clean %>%
  left_join(entry_year_lookup, by = "player_name")

recovered <- first_year_complete %>%
  left_join(first_year_trends, by = "player_name") %>%
  filter(!is.na(first_year_trends) & first_year_trends < first_year_complete)

cat("Players with entry_year recovered earlier via trends file:", nrow(recovered), "\n")
cat("Players left_truncated (entry_year == trends file start,", TRENDS_START_YEAR, "):",
    sum(entry_year_lookup$left_truncated), "\n")
```

Spot check a few recovered players against known PGA Tour debuts:

```{r entry-year-spot-check}
recovered %>%
  left_join(entry_year_lookup, by = "player_name") %>%
  arrange(entry_year) %>%
  select(player_name, first_year_complete, entry_year) %>%
  head(10) %>%
  print()
```

---

# 4. Engineer Variables — df_complete

## 4.1 Career tenure and cohort variables

`entry_year` (section 3.4) is the true, cross-referenced tour-debut year and
is the right time origin for tenure. `cohort` is split into two purpose-built
variables rather than one: `entry_cohort` (decade of tour entry) is the
right lens for career-stage/tenure questions (survival analysis, time-to-
obsolescence); `birth_cohort` (decade of birth) isolates generational
effects (equipment/training exposure by birth era) without conflating a
young rookie with a late-blooming veteran who happened to enter the tour in
the same decade.

```{r career-vars}
df_clean <- df_clean %>%
  mutate(
    # Tenure on tour, from the cross-referenced true entry year
    years_on_tour = year - entry_year,

    # Entry cohort: decade of true tour entry (career-stage lens)
    entry_cohort = case_when(
      entry_year < 2000                      ~ "pre-2000",
      entry_year >= 2000 & entry_year < 2010  ~ "2000–2009",
      entry_year >= 2010 & entry_year < 2020  ~ "2010–2019",
      entry_year >= 2020                      ~ "2020+",
      TRUE                                    ~ NA_character_
    ) %>% factor(levels = c("pre-2000", "2000–2009", "2010–2019", "2020+")),

    # Birth cohort: decade of birth (generational lens)
    birth_cohort = case_when(
      birth_year < 1970                     ~ "Pre-1970",
      birth_year >= 1970 & birth_year < 1980 ~ "1970s",
      birth_year >= 1980 & birth_year < 1990 ~ "1980s",
      birth_year >= 1990 & birth_year < 2000 ~ "1990s",
      birth_year >= 2000                     ~ "2000s+",
      TRUE                                   ~ NA_character_
    ) %>% factor(levels = c("Pre-1970", "1970s", "1980s", "1990s", "2000s+"))
  )

df_clean %>%
  count(entry_cohort) %>%
  print()

df_clean %>%
  count(birth_cohort) %>%
  print()
```

## 4.2 Field-relative metrics (within-year centering)

These express each player's stats relative to the season average — removing
the secular trend so we can compare players across eras.

```{r field-relative}
df_clean <- df_clean %>%
  group_by(year) %>%
  mutate(
    dist_vs_field  = distance   - mean(distance,   na.rm = TRUE),
    acc_vs_field   = accuracy   - mean(accuracy,   na.rm = TRUE),
    speed_vs_field = ball_speed - mean(ball_speed, na.rm = TRUE),
    chs_vs_field   = chs        - mean(chs,        na.rm = TRUE)
  ) %>%
  ungroup()
```

## 4.3 Mundlak decomposition variables

The Mundlak approach separates the *between-player* effect of distance
(do longer hitters systematically differ from shorter hitters?) from the
*within-player* effect (does a player improve when their own distance
increases?). This is critical for correctly interpreting panel models.

```{r mundlak}
df_clean <- df_clean %>%
  group_by(player_name) %>%
  mutate(
    # Career-average distance for each player (between-player component)
    dist_player_mean = mean(distance, na.rm = TRUE),

    # Deviation from own career average (within-player component)
    dist_within = distance - dist_player_mean
  ) %>%
  ungroup()
```

## 4.4 Distance quintiles (within each year)

Players ranked into five equal groups within each season. Q1 = shortest,
Q5 = longest.

```{r quintiles}
df_clean <- df_clean %>%
  group_by(year) %>%
  mutate(
    dist_quintile = ntile(distance, 5)
  ) %>%
  ungroup()

# Verify distribution is roughly even
df_clean %>%
  count(dist_quintile) %>%
  mutate(pct = round(n / sum(n) * 100, 1)) %>%
  print()
```

## 4.5 Player archetype (quadrant classification)

Based on whether a player is above or below the season average in both
distance and accuracy. Simple, interpretable, and robust to year-over-year
changes in absolute values.

```{r archetypes}
df_clean <- df_clean %>%
  mutate(
    archetype = case_when(
      dist_vs_field >= 0 & acc_vs_field >= 0 ~ "Long / Accurate",
      dist_vs_field >= 0 & acc_vs_field <  0 ~ "Long / Inaccurate",
      dist_vs_field <  0 & acc_vs_field >= 0 ~ "Short / Accurate",
      dist_vs_field <  0 & acc_vs_field <  0 ~ "Short / Inaccurate",
      TRUE ~ NA_character_
    ) %>% factor(levels = c(
      "Long / Accurate", "Long / Inaccurate",
      "Short / Accurate", "Short / Inaccurate"
    ))
  )

df_clean %>%
  count(archetype) %>%
  mutate(pct = round(n / sum(n) * 100, 1)) %>%
  print()
```

## 4.6 Career entry distance (strokes-gained era)

Each player's distance in their first *strokes-gained-era* observed season
(`sg_entry_year`) — distinct from the cross-referenced `entry_year` in
section 3.4, since ball speed/CHS only exist within `df_complete`'s 2004+
window. Used in survival analysis and rising floor analyses to ask "what did
you arrive with, once we could measure speed?"

```{r entry-distance}
entry_stats <- df_clean %>%
  group_by(player_name) %>%
  slice_min(year, n = 1, with_ties = FALSE) %>%
  select(player_name, dist_entry = distance, speed_entry = ball_speed,
         chs_entry = chs, sg_entry_year = year) %>%
  ungroup()

df_clean <- df_clean %>%
  left_join(entry_stats, by = "player_name")

# Quick check
df_clean %>%
  select(player_name, year, distance, dist_entry, sg_entry_year) %>%
  filter(player_name == first(player_name)) %>%
  print()
```

## 4.7 Distance quartile at career entry

Used in survival analysis as a time-fixed predictor — what distance tier
did a player arrive in (within the strokes-gained era)?

```{r entry-quartile}
# Quartile within entry year's field
entry_quartiles <- df_clean %>%
  filter(year == sg_entry_year) %>%
  group_by(year) %>%
  mutate(dist_entry_quartile = ntile(distance, 4)) %>%
  ungroup() %>%
  select(player_name, dist_entry_quartile)

df_clean <- df_clean %>%
  left_join(entry_quartiles, by = "player_name")

df_clean %>%
  count(dist_entry_quartile) %>%
  print()
```

---

# 5. Clean and Engineer df_trends

## 5.1 Standardise column names

```{r trends-clean}
df_trends <- df_trends_raw %>%
  rename(player_name = player) %>%
  mutate(
    player_name = as.character(player_name),
    year        = as.integer(year)
  )
```

## 5.2 Minimum seasons filter

```{r trends-min-seasons}
seasons_trends <- df_trends %>%
  count(player_name, name = "n_seasons")

df_trends <- df_trends %>%
  left_join(seasons_trends, by = "player_name") %>%
  mutate(multi_season = n_seasons >= MIN_SEASONS)

cat("Players with 1 season only: ",
    sum(seasons_trends$n_seasons < MIN_SEASONS), "\n")
cat("Players with 2+ seasons:    ",
    sum(seasons_trends$n_seasons >= MIN_SEASONS), "\n")
```

## 5.3 Verify and standardise field-relative metrics

The raw file already contains `relative_distance` and `relative_accuracy`
(player value minus season mean). We verify these match our own calculation
and rename to match the `df_main` convention.

```{r trends-relative}
# Recalculate to verify
df_trends <- df_trends %>%
  group_by(year) %>%
  mutate(
    dist_vs_field_check = distance - mean(distance, na.rm = TRUE),
    acc_vs_field_check  = accuracy - mean(accuracy, na.rm = TRUE)
  ) %>%
  ungroup()

# Compare to existing columns (should be near zero difference)
df_trends %>%
  summarise(
    dist_diff_max = max(abs(relative_distance - dist_vs_field_check), na.rm = TRUE),
    acc_diff_max  = max(abs(relative_accuracy - acc_vs_field_check),  na.rm = TRUE)
  ) %>%
  print()

# Rename to standard convention and drop check columns
df_trends <- df_trends %>%
  rename(
    dist_vs_field = relative_distance,
    acc_vs_field  = relative_accuracy
  ) %>%
  select(-dist_vs_field_check, -acc_vs_field_check)
```

## 5.4 Add cohort variables

```{r trends-cohort}
# This file already goes back to 1987, so its own first-observed-year is
# already a good entry-year estimate for players who don't reach df_complete
# (no separate cross-referencing needed, unlike df_main's entry_cohort).
first_year_trends <- df_trends %>%
  group_by(player_name) %>%
  summarise(first_year = min(year)) %>%
  ungroup()

df_trends <- df_trends %>%
  left_join(first_year_trends, by = "player_name") %>%
  mutate(
    entry_cohort = case_when(
      first_year < 2000              ~ "pre-2000",
      first_year >= 2000 & first_year < 2010 ~ "2000–2009",
      first_year >= 2010 & first_year < 2020 ~ "2010–2019",
      first_year >= 2020             ~ "2020+",
      TRUE                           ~ NA_character_
    ) %>% factor(levels = c("pre-2000", "2000–2009", "2010–2019", "2020+"))
  )

# Enrich with birth_cohort where the player also appears in df_complete.
# Coverage is partial: df_complete only covers 793 of df_trends's 1184 players.
birth_lookup <- df_clean %>%
  distinct(player_name, birth_cohort)

df_trends <- df_trends %>%
  left_join(birth_lookup, by = "player_name")

cat("Players with birth_cohort available (matched to df_complete):",
    n_distinct(df_trends$player_name[!is.na(df_trends$birth_cohort)]), "\n")
cat("Total players in df_trends:                                 ",
    n_distinct(df_trends$player_name), "\n")
```

---

# 6. Final Checks

## 6.1 df_main summary

```{r final-check-main}
cat("=== df_main ===\n")
cat("Rows:              ", nrow(df_clean), "\n")
cat("Unique players:    ", n_distinct(df_clean$player_name), "\n")
cat("Year range:        ", min(df_clean$year), "–", max(df_clean$year), "\n")
cat("Engineered columns:\n")
cat(" ", paste(c("entry_year", "left_truncated", "years_on_tour",
                 "entry_cohort", "birth_cohort", "dist_vs_field", "acc_vs_field",
                 "speed_vs_field", "chs_vs_field", "dist_player_mean",
                 "dist_within", "dist_quintile", "archetype",
                 "dist_entry", "sg_entry_year", "dist_entry_quartile"), collapse = "\n  "), "\n")

# Spot check a well-known player if present
if ("Tiger Woods" %in% df_clean$player_name) {
  df_clean %>%
    filter(player_name == "Tiger Woods") %>%
    select(year, distance, dist_vs_field, dist_quintile, archetype,
           entry_year, entry_cohort, birth_cohort, left_truncated) %>%
    print()
}
```

## 6.2 df_trends summary

```{r final-check-trends}
cat("=== df_trends ===\n")
cat("Rows:              ", nrow(df_trends), "\n")
cat("Unique players:    ", n_distinct(df_trends$player_name), "\n")
cat("Year range:        ", min(df_trends$year), "–", max(df_trends$year), "\n")
```

---

# 7. Write Outputs

```{r write-outputs}
saveRDS(df_clean,  here(PATH_PROCESSED, "df_main.rds"))
saveRDS(df_trends, here(PATH_PROCESSED, "df_trends.rds"))

cat("Outputs written to data/processed/:\n")
cat("  df_main.rds  — ", nrow(df_clean),  "rows,",
    n_distinct(df_clean$player_name),  "players\n")
cat("  df_trends.rds —", nrow(df_trends), "rows,",
    n_distinct(df_trends$player_name), "players\n")
```

---

# 8. Session Info

```{r session-info}
sessionInfo()
```