---
title: "00 — Data Engineering"
subtitle: "PGA Tour Distance & Speed Analytics"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    df_print: paged
---

# Overview

This notebook is the single source of truth for all data cleaning and variable
engineering in this project. It reads the two raw datasets, applies cleaning
decisions, engineers all derived variables, and writes processed outputs to
`data/processed/`. Every subsequent analysis notebook loads from those outputs
rather than re-processing the raw data.

**Inputs:**
- `data/raw/df_complete.csv` — Player-season records, 2004–2025
- `data/raw/trends_acc_dist_1987_to_2025.csv` — Distance and accuracy, 1987–2025

**Outputs:**
- `data/processed/df_main.rds` — Cleaned and engineered df_complete
- `data/processed/df_trends.rds` — Cleaned df_trends

**Note on survival data:** The survival analysis dataset (`df_survival.rds`) is
constructed separately in `00b_survival_prep.Rmd` once we have sufficient
familiarity with career gap patterns in the data to make defensible censoring
decisions.

---

# 1. Setup

```{r setup, message=FALSE, warning=FALSE}
# Core
library(tidyverse)
library(here)

# Utilities
library(broom)

# Source shared config (paths, constants, regulatory events, plot theme)
source(here::here("config.R"))

# Reproducibility
set.seed(2025)

# Global knitr options
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 10,
  fig.height = 6
)
```

---

# 2. Load Raw Data

```{r load-data}
df_raw <- read_csv(
  here(PATH_RAW, "df_complete.csv"),
  show_col_types = FALSE
)

df_trends_raw <- read_csv(
  here(PATH_RAW, "trends_acc_dist_1987_to_2025.csv"),
  show_col_types = FALSE
)
```

## 2.1 Initial inspection — df_complete

We rely on `player_name` + `year` as the unique key for this dataset in every
downstream `left_join`. Confirm that holds before trusting any join result.

```{r key-check-complete}
n_dupes_complete <- df_raw %>%
  count(player_name, year) %>%
  filter(n > 1) %>%
  nrow()

cat("Rows:                    ", nrow(df_raw), "\n")
cat("Distinct player+year keys:", n_distinct(df_raw$player_name, df_raw$year), "\n")
cat("Duplicate player+year keys:", n_dupes_complete, "\n")

stopifnot(n_dupes_complete == 0)
```

```{r inspect-complete}
glimpse(df_raw)
```

```{r inspect-complete-summary}
cat("Rows:           ", nrow(df_raw), "\n")
cat("Unique players: ", n_distinct(df_raw$player_name), "\n")
cat("Year range:     ", min(df_raw$year), "–", max(df_raw$year), "\n")
cat("\nNA counts:\n")
df_raw %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "column", values_to = "n_missing") %>%
  filter(n_missing > 0) %>%
  print()
```

Sanity-check that key measures fall within plausible bounds for PGA Tour
play. This is a coarse guard against data-entry errors, not a claim about
true physical limits — values outside these ranges are reported, not
dropped.

```{r range-check-complete}
df_raw %>%
  summarise(
    accuracy_out_of_range = sum(accuracy < 0 | accuracy > 100, na.rm = TRUE),
    distance_out_of_range = sum(distance < 200 | distance > 350, na.rm = TRUE),
    ball_speed_out_of_range = sum(ball_speed < 140 | ball_speed > 200, na.rm = TRUE),
    chs_out_of_range = sum(chs < 90 | chs > 140, na.rm = TRUE),
    carry_out_of_range = sum(carry_distance < 200 | carry_distance > 320, na.rm = TRUE)
  ) %>%
  print()
```

## 2.2 Initial inspection — df_trends

```{r key-check-trends}
n_dupes_trends <- df_trends_raw %>%
  count(player, year) %>%
  filter(n > 1) %>%
  nrow()

cat("Rows:                    ", nrow(df_trends_raw), "\n")
cat("Distinct player+year keys:", n_distinct(df_trends_raw$player, df_trends_raw$year), "\n")
cat("Duplicate player+year keys:", n_dupes_trends, "\n")

stopifnot(n_dupes_trends == 0)
```

```{r inspect-trends}
glimpse(df_trends_raw)
```

```{r inspect-trends-summary}
cat("Rows:           ", nrow(df_trends_raw), "\n")
cat("Unique players: ", n_distinct(df_trends_raw$player), "\n")
cat("Year range:     ", min(df_trends_raw$year), "–", max(df_trends_raw$year), "\n")
cat("\nNA counts:\n")
df_trends_raw %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "column", values_to = "n_missing") %>%
  filter(n_missing > 0) %>%
  print()
```

```{r range-check-trends}
df_trends_raw %>%
  summarise(
    accuracy_out_of_range = sum(accuracy < 0 | accuracy > 100, na.rm = TRUE),
    distance_out_of_range = sum(distance < 200 | distance > 350, na.rm = TRUE)
  ) %>%
  print()
```

---

# 3. Clean df_complete

## 3.1 Column names and types

```{r clean-colnames}
df_clean <- df_raw %>%
  # Fix column name with space
  rename(rookie_year = `rookie year`) %>%
  # Ensure player_name is character, year is integer
  mutate(
    player_name = as.character(player_name),
    year        = as.integer(year)
  )

# Confirm rename worked
names(df_clean)
```

## 3.2 Handle the pre-2007 speed data gap

Ball speed, CHS, and carry distance have ~594 NAs concentrated in 2004–2006,
before these metrics were tracked by ShotLink. We do not impute these values.
Instead we document the gap and note the effective start year for speed analyses.

```{r speed-na-check}
df_clean %>%
  group_by(year) %>%
  summarise(
    n             = n(),
    na_ball_speed = sum(is.na(ball_speed)),
    na_chs        = sum(is.na(chs)),
    na_carry      = sum(is.na(carry_distance))
  ) %>%
  filter(na_ball_speed > 0) %>%
  print()
```

The `SPEED_ERA_START` constant in `config.R` is set to `2007`. Any analysis
involving ball speed or CHS should filter to `year >= SPEED_ERA_START`.

## 3.2b Missing rookie_year / turned_pro

As with the speed metrics, we do not impute these — we document the gap and
carry the NAs through into any variable derived from them.

```{r rookie-year-na-check}
missing_rookie_year <- df_clean %>%
  filter(is.na(rookie_year))

cat("Rows missing rookie_year:   ", nrow(missing_rookie_year), "\n")
cat("Players missing rookie_year:", n_distinct(missing_rookie_year$player_name), "\n")
cat("Rows missing turned_pro:    ", sum(is.na(df_clean$turned_pro)), "\n")
```

Because `rookie_year` is missing for all of a given player's rows rather
than sporadically within a player's history, `years_on_tour` (section 4.1)
will be NA for every row belonging to these players. Similarly, the single
row with missing `turned_pro` will have an NA `cohort` (section 4.1). No
imputation is applied; any analysis that depends on `years_on_tour` or
`cohort` should be aware these players/rows will be excluded via NA
propagation rather than explicitly filtered.

## 3.3 Minimum seasons filter

Players appearing in fewer than `MIN_SEASONS` (2) seasons are excluded from
longitudinal analyses. We flag rather than drop them here, preserving the
option to include single-season players in purely cross-sectional work.

```{r min-seasons}
seasons_per_player <- df_clean %>%
  count(player_name, name = "n_seasons")

df_clean <- df_clean %>%
  left_join(seasons_per_player, by = "player_name") %>%
  mutate(multi_season = n_seasons >= MIN_SEASONS)

cat("Players with 1 season only: ",
    sum(seasons_per_player$n_seasons < MIN_SEASONS), "\n")
cat("Players with 2+ seasons:    ",
    sum(seasons_per_player$n_seasons >= MIN_SEASONS), "\n")
```

---

# 4. Engineer Variables — df_complete

## 4.1 Career and cohort variables

```{r career-vars}
df_clean <- df_clean %>%
  mutate(
    # Years on tour (from rookie year)
    years_on_tour = year - rookie_year,

    # Entry cohort by decade turned pro
    cohort = case_when(
      turned_pro < 2000              ~ "pre-2000",
      turned_pro >= 2000 & turned_pro < 2010 ~ "2000–2009",
      turned_pro >= 2010 & turned_pro < 2020 ~ "2010–2019",
      turned_pro >= 2020             ~ "2020+",
      TRUE                           ~ NA_character_
    ) %>% factor(levels = c("pre-2000", "2000–2009", "2010–2019", "2020+"))
  )

df_clean %>%
  count(cohort) %>%
  print()
```

## 4.2 Field-relative metrics (within-year centering)

These express each player's stats relative to the season average — removing
the secular trend so we can compare players across eras.

```{r field-relative}
df_clean <- df_clean %>%
  group_by(year) %>%
  mutate(
    dist_vs_field  = distance   - mean(distance,   na.rm = TRUE),
    acc_vs_field   = accuracy   - mean(accuracy,   na.rm = TRUE),
    speed_vs_field = ball_speed - mean(ball_speed, na.rm = TRUE),
    chs_vs_field   = chs        - mean(chs,        na.rm = TRUE)
  ) %>%
  ungroup()
```

## 4.3 Mundlak decomposition variables

The Mundlak approach separates the *between-player* effect of distance
(do longer hitters systematically differ from shorter hitters?) from the
*within-player* effect (does a player improve when their own distance
increases?). This is critical for correctly interpreting panel models.

```{r mundlak}
df_clean <- df_clean %>%
  group_by(player_name) %>%
  mutate(
    # Career-average distance for each player (between-player component)
    dist_player_mean = mean(distance, na.rm = TRUE),

    # Deviation from own career average (within-player component)
    dist_within = distance - dist_player_mean
  ) %>%
  ungroup()
```

## 4.4 Distance quintiles (within each year)

Players ranked into five equal groups within each season. Q1 = shortest,
Q5 = longest.

```{r quintiles}
df_clean <- df_clean %>%
  group_by(year) %>%
  mutate(
    dist_quintile = ntile(distance, 5)
  ) %>%
  ungroup()

# Verify distribution is roughly even
df_clean %>%
  count(dist_quintile) %>%
  mutate(pct = round(n / sum(n) * 100, 1)) %>%
  print()
```

## 4.5 Player archetype (quadrant classification)

Based on whether a player is above or below the season average in both
distance and accuracy. Simple, interpretable, and robust to year-over-year
changes in absolute values.

```{r archetypes}
df_clean <- df_clean %>%
  mutate(
    archetype = case_when(
      dist_vs_field >= 0 & acc_vs_field >= 0 ~ "Long / Accurate",
      dist_vs_field >= 0 & acc_vs_field <  0 ~ "Long / Inaccurate",
      dist_vs_field <  0 & acc_vs_field >= 0 ~ "Short / Accurate",
      dist_vs_field <  0 & acc_vs_field <  0 ~ "Short / Inaccurate",
      TRUE ~ NA_character_
    ) %>% factor(levels = c(
      "Long / Accurate", "Long / Inaccurate",
      "Short / Accurate", "Short / Inaccurate"
    ))
  )

df_clean %>%
  count(archetype) %>%
  mutate(pct = round(n / sum(n) * 100, 1)) %>%
  print()
```

## 4.6 Career entry distance

Each player's distance in their first observed season. Used in survival
analysis and rising floor analyses to ask "what did you arrive with?"

```{r entry-distance}
entry_stats <- df_clean %>%
  group_by(player_name) %>%
  slice_min(year, n = 1, with_ties = FALSE) %>%
  select(player_name, dist_entry = distance, speed_entry = ball_speed,
         chs_entry = chs, entry_year = year) %>%
  ungroup()

df_clean <- df_clean %>%
  left_join(entry_stats, by = "player_name")

# Quick check
df_clean %>%
  select(player_name, year, distance, dist_entry, entry_year) %>%
  filter(player_name == first(player_name)) %>%
  print()
```

## 4.7 Distance quartile at career entry

Used in survival analysis as a time-fixed predictor — what distance
tier did a player arrive in?

```{r entry-quartile}
# Quartile within entry year's field
entry_quartiles <- df_clean %>%
  filter(year == entry_year) %>%
  group_by(year) %>%
  mutate(dist_entry_quartile = ntile(distance, 4)) %>%
  ungroup() %>%
  select(player_name, dist_entry_quartile)

df_clean <- df_clean %>%
  left_join(entry_quartiles, by = "player_name")

df_clean %>%
  count(dist_entry_quartile) %>%
  print()
```

---

# 5. Clean and Engineer df_trends

## 5.1 Standardise column names

```{r trends-clean}
df_trends <- df_trends_raw %>%
  rename(player_name = player) %>%
  mutate(
    player_name = as.character(player_name),
    year        = as.integer(year)
  )
```

## 5.2 Minimum seasons filter

```{r trends-min-seasons}
seasons_trends <- df_trends %>%
  count(player_name, name = "n_seasons")

df_trends <- df_trends %>%
  left_join(seasons_trends, by = "player_name") %>%
  mutate(multi_season = n_seasons >= MIN_SEASONS)

cat("Players with 1 season only: ",
    sum(seasons_trends$n_seasons < MIN_SEASONS), "\n")
cat("Players with 2+ seasons:    ",
    sum(seasons_trends$n_seasons >= MIN_SEASONS), "\n")
```

## 5.3 Verify and standardise field-relative metrics

The raw file already contains `relative_distance` and `relative_accuracy`
(player value minus season mean). We verify these match our own calculation
and rename to match the `df_main` convention.

```{r trends-relative}
# Recalculate to verify
df_trends <- df_trends %>%
  group_by(year) %>%
  mutate(
    dist_vs_field_check = distance - mean(distance, na.rm = TRUE),
    acc_vs_field_check  = accuracy - mean(accuracy, na.rm = TRUE)
  ) %>%
  ungroup()

# Compare to existing columns (should be near zero difference)
df_trends %>%
  summarise(
    dist_diff_max = max(abs(relative_distance - dist_vs_field_check), na.rm = TRUE),
    acc_diff_max  = max(abs(relative_accuracy - acc_vs_field_check),  na.rm = TRUE)
  ) %>%
  print()

# Rename to standard convention and drop check columns
df_trends <- df_trends %>%
  rename(
    dist_vs_field = relative_distance,
    acc_vs_field  = relative_accuracy
  ) %>%
  select(-dist_vs_field_check, -acc_vs_field_check)
```

## 5.4 Add cohort variable

```{r trends-cohort}
# For trends file we don't have turned_pro directly, so derive from
# first observed year as a proxy for cohort grouping
first_year_trends <- df_trends %>%
  group_by(player_name) %>%
  summarise(first_year = min(year)) %>%
  ungroup()

df_trends <- df_trends %>%
  left_join(first_year_trends, by = "player_name") %>%
  mutate(
    cohort = case_when(
      first_year < 2000              ~ "pre-2000",
      first_year >= 2000 & first_year < 2010 ~ "2000–2009",
      first_year >= 2010 & first_year < 2020 ~ "2010–2019",
      first_year >= 2020             ~ "2020+",
      TRUE                           ~ NA_character_
    ) %>% factor(levels = c("pre-2000", "2000–2009", "2010–2019", "2020+"))
  )
```

---

# 6. Final Checks

## 6.1 df_main summary

```{r final-check-main}
cat("=== df_main ===\n")
cat("Rows:              ", nrow(df_clean), "\n")
cat("Unique players:    ", n_distinct(df_clean$player_name), "\n")
cat("Year range:        ", min(df_clean$year), "–", max(df_clean$year), "\n")
cat("Engineered columns:\n")
cat(" ", paste(c("years_on_tour", "cohort", "dist_vs_field", "acc_vs_field",
                 "speed_vs_field", "chs_vs_field", "dist_player_mean",
                 "dist_within", "dist_quintile", "archetype",
                 "dist_entry", "dist_entry_quartile"), collapse = "\n  "), "\n")

# Spot check a well-known player if present
if ("Tiger Woods" %in% df_clean$player_name) {
  df_clean %>%
    filter(player_name == "Tiger Woods") %>%
    select(year, distance, dist_vs_field, dist_quintile, archetype, cohort) %>%
    print()
}
```

## 6.2 df_trends summary

```{r final-check-trends}
cat("=== df_trends ===\n")
cat("Rows:              ", nrow(df_trends), "\n")
cat("Unique players:    ", n_distinct(df_trends$player_name), "\n")
cat("Year range:        ", min(df_trends$year), "–", max(df_trends$year), "\n")
```

---

# 7. Write Outputs

```{r write-outputs}
saveRDS(df_clean,  here(PATH_PROCESSED, "df_main.rds"))
saveRDS(df_trends, here(PATH_PROCESSED, "df_trends.rds"))

cat("Outputs written to data/processed/:\n")
cat("  df_main.rds  — ", nrow(df_clean),  "rows,",
    n_distinct(df_clean$player_name),  "players\n")
cat("  df_trends.rds —", nrow(df_trends), "rows,",
    n_distinct(df_trends$player_name), "players\n")
```

---

# 8. Session Info

```{r session-info}
sessionInfo()
```