---
title: "00b — Survival Data Preparation"
subtitle: "PGA Tour Distance & Speed Analytics"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    df_print: paged
---

# Overview

This notebook builds the survival-analysis dataset from `df_main.rds`. PGA
Tour careers are not a clean rectangular panel — 35% of players miss at
least one season, and in 165 individual cases a player missed 2+
*consecutive* seasons and later returned (conditional status, injury, tour
crossover, etc.). A single fixed rule like "2+ consecutive seasons absent =
career ended" was checked directly against the data and found to
misclassify well over 100 still-active careers.

So this notebook does **not** decide what counts as "career ended." It only
reconstructs the underlying structure — each player's distinct spells of
consecutive seasons, and the gaps between them — and leaves the actual
censoring/event threshold as an explicit, testable choice for
`07_survival_analysis.Rmd` (e.g. comparing 2/3/4-year gap thresholds as a
named sensitivity check, per `docs/project_context.md`).

**Input:** `data/processed/df_main.rds`
**Output:** `data/processed/df_survival.rds`

---

# 1. Setup

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(here)

source(here::here("config.R"))

set.seed(2025)

knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 10,
  fig.height = 6
)
```

---

# 2. Load Processed Data

```{r load-data}
df_main <- readRDS(here(PATH_PROCESSED, "df_main.rds"))

DATA_END_YEAR <- max(df_main$year)

cat("Rows:          ", nrow(df_main), "\n")
cat("Unique players:", n_distinct(df_main$player_name), "\n")
cat("Data end year: ", DATA_END_YEAR, "\n")
```

---

# 3. Reconstruct Spells

A "spell" is a run of consecutive observed seasons for a player. A new spell
starts whenever there's a gap of more than one year since the player's
previous observed season.

```{r spell-ids}
spells <- df_main %>%
  distinct(player_name, year) %>%
  arrange(player_name, year) %>%
  group_by(player_name) %>%
  mutate(
    gap_flag = coalesce(year - lag(year) > 1, FALSE),
    spell_id = cumsum(gap_flag) + 1
  ) %>%
  ungroup()

cat("Distribution of spells-per-player:\n")
spells %>%
  count(player_name, name = "n_years") %>%
  count(player_name) %>%
  nrow()

spells %>%
  group_by(player_name) %>%
  summarise(n_spells = max(spell_id), .groups = "drop") %>%
  count(n_spells) %>%
  print()
```

## 3.1 Spell-level summary

```{r spell-summary}
spell_summary <- spells %>%
  group_by(player_name, spell_id) %>%
  summarise(
    spell_start_year = min(year),
    spell_end_year   = max(year),
    .groups = "drop"
  ) %>%
  group_by(player_name) %>%
  arrange(spell_id, .by_group = TRUE) %>%
  mutate(
    gap_before        = spell_start_year - lag(spell_end_year) - 1,
    gap_after         = lead(spell_start_year) - spell_end_year - 1,
    is_final_spell    = spell_id == max(spell_id),
    final_gap_length  = if_else(is_final_spell, DATA_END_YEAR - spell_end_year, NA_integer_)
  ) %>%
  ungroup()

# Every player should have exactly one final spell
final_spell_counts <- spell_summary %>%
  filter(is_final_spell) %>%
  count(player_name) %>%
  filter(n != 1)

cat("Players without exactly one final spell (should be 0):", nrow(final_spell_counts), "\n")
stopifnot(nrow(final_spell_counts) == 0)
```

## 3.2 Spot check — Arjun Atwal

Known pattern: seasons 2004–2006, 2011–2012, 2020 (two multi-year gaps,
including a 7-year gap before returning in 2020).

```{r spot-check-atwal}
spell_summary %>%
  filter(player_name == "Arjun Atwal") %>%
  print()
```

---

# 4. Build Survival Dataset

Join spell structure back onto each player-year row, and add the start-stop
time axis (`time_start`/`time_stop`) using calendar tenure from the
cross-referenced `entry_year` (not `rookie_year` — see
`00_data_engineering.Rmd` sections 3.2b/3.4/4.1 for why).

```{r build-survival}
spells_with_id <- spells %>% select(player_name, year, spell_id)

df_survival <- df_main %>%
  left_join(spells_with_id, by = c("player_name", "year")) %>%
  left_join(spell_summary, by = c("player_name", "spell_id")) %>%
  mutate(
    time_start = year - entry_year,
    time_stop  = time_start + 1
  ) %>%
  select(
    player_name, year, entry_year, left_truncated, entry_cohort,
    spell_id, spell_start_year, spell_end_year,
    gap_before, gap_after, is_final_spell, final_gap_length,
    time_start, time_stop,
    dist_entry_quartile, dist_vs_field, sg_total, age
  )

cat("Rows in df_survival:", nrow(df_survival), "\n")
cat("Rows in df_main:    ", nrow(df_main), "\n")
stopifnot(nrow(df_survival) == nrow(df_main))
```

---

# 5. Final Checks

```{r final-checks}
cat("=== df_survival ===\n")
cat("Rows:              ", nrow(df_survival), "\n")
cat("Unique players:    ", n_distinct(df_survival$player_name), "\n")
cat("Year range:        ", min(df_survival$year), "–", max(df_survival$year), "\n")

cat("\nleft_truncated players:", n_distinct(df_survival$player_name[df_survival$left_truncated]), "\n")

cat("\nfinal_gap_length distribution (final-spell rows only):\n")
df_survival %>%
  filter(is_final_spell) %>%
  count(final_gap_length) %>%
  arrange(final_gap_length) %>%
  print()

# Spot check a well-known player if present
if ("Tiger Woods" %in% df_survival$player_name) {
  df_survival %>%
    filter(player_name == "Tiger Woods") %>%
    select(year, entry_year, spell_id, gap_before, gap_after, time_start, time_stop) %>%
    print()
}
```

---

# 6. Write Output

```{r write-output}
saveRDS(df_survival, here(PATH_PROCESSED, "df_survival.rds"))

cat("Output written to data/processed/df_survival.rds —",
    nrow(df_survival), "rows,", n_distinct(df_survival$player_name), "players\n")
```

---

# 7. Session Info

```{r session-info}
sessionInfo()
```
