---
title: "02 — Regulatory Inflection Points"
subtitle: "PGA Tour Distance & Speed Analytics"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    df_print: paged
---

# Overview

**Question:** Did equipment and rule changes produce statistically
detectable changes in distance and accuracy trends?

**Data:** `df_trends` aggregated to a season-mean time series (one row per
year), not the player-level panel used in Project 1.

**A design note before starting:** `config.R`'s `regulatory_events` lists 7
events, but 2003 (COR Limits) and 2004 (Club Size/MOI Limits) are one year
apart — annual data can't separately identify two breakpoints that close
together (confirmed below). We merge them into one combined event,
**"2003-04 COR/MOI Limits,"** rather than just noting the problem and
moving on. The resulting 6 events are also labeled **"Equipment advance"**
(1992 Oversized Drivers, 1996 Titanium Drivers, 2000 Pro V1) or
**"Regulatory restriction"** (2003-04 COR/MOI Limits, 2010 Groove Rule, 2022
Driver Length Limit) — a distinction that organizes the summary around this
project's actual thesis: did the restrictive rules slow anything down, or
did the enabling technology do the real work?

We also don't assume effects are instant. A rule change might take 1-3
years to show up as players adopt new equipment — so wherever a breakpoint
or intervention date is estimated, we report the *lag* between the nominal
regulatory year and the year the data actually shows a shift.

---

# 1. Setup

```{r setup, message=FALSE, warning=FALSE}
# Core
library(tidyverse)
library(here)

# Modeling
library(segmented)   # segmented regression, known + selected breakpoints (section 3)
library(changepoint) # PELT changepoint detection (section 4)
library(strucchange) # Bai-Perron breakpoints, CUSUM (section 4)

# Utilities
library(broom)
library(scales)

source(here::here("config.R"))

# segmented/changepoint/strucchange pull in MASS as a dependency, which masks
# dplyr::select() - force dplyr's version to win for the rest of this notebook.
select <- dplyr::select

set.seed(2025)

knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 10,
  fig.height = 6
)
```

---

# 2. Load and Aggregate Data

```{r load-data}
df_trends <- readRDS(here(PATH_PROCESSED, "df_trends.rds"))

season_means <- df_trends %>%
  group_by(year) %>%
  summarise(
    n             = n(),
    mean_distance = mean(distance, na.rm = TRUE),
    mean_accuracy = mean(accuracy, na.rm = TRUE),
    .groups       = "drop"
  ) %>%
  arrange(year)

DATA_END_YEAR <- max(season_means$year)

cat("Years:      ", nrow(season_means), "\n")
cat("Year range: ", min(season_means$year), "-", DATA_END_YEAR, "\n")
```

## 2.1 Build the working event table

Merge 2003 COR Limits + 2004 Club Size/MOI Limits into one event (see
Overview), and label each event's category.

```{r build-events}
events <- regulatory_events %>%
  mutate(
    year  = if_else(year == 2004, 2003, year),
    label = if_else(year == 2003, "COR/MOI Limits", label)
  ) %>%
  distinct(year, label) %>%
  mutate(
    category = if_else(year %in% c(1992, 1996, 2000), "Equipment advance", "Regulatory restriction")
  ) %>%
  arrange(year)

events %>% print()
```

---

# 3. Annotated Baseline Trend

The visual starting point before any modeling — season-mean distance with
each event marked, so the reader can eyeball candidate inflection points
themselves before we quantify anything.

```{r baseline-trend}
p_baseline <- ggplot(season_means, aes(year, mean_distance)) +
  geom_vline(data = events, aes(xintercept = year, color = category),
             linetype = "dashed", linewidth = 0.6) +
  geom_line(color = COL_PRIMARY, linewidth = 1) +
  geom_point(color = COL_PRIMARY, size = 1.3) +
  scale_color_manual(values = c(
    "Equipment advance"      = COL_SECONDARY,
    "Regulatory restriction" = COL_ACCENT
  )) +
  labs(
    title    = "Season-Mean Driving Distance with Regulatory Events",
    subtitle = "Dashed lines mark regulatory events (table below)",
    x = NULL, y = "Mean Driving Distance (yards)", color = NULL
  ) +
  THEME_BASE

p_baseline

events %>% select(year, label, category) %>% print()
```

---

# 4. Segmented Regression With Known Breakpoints

We first tried supplying all 6 event years as candidate breakpoints to
`segmented::segmented()` simultaneously — the honest result was that the
model doesn't converge cleanly and several breakpoints come back with
enormous standard errors (20+ years), meaning the ~39-year series simply
can't support 6 independent breakpoints at once. Rather than report an
unreliable 6-breakpoint model, we use `segmented::selgmented()`, which lets
BIC pick how many breakpoints the data actually supports, then compare
those data-supported breakpoints to the nominal event years.

```{r segmented-naive-attempt}
fit_lm <- lm(mean_distance ~ year, data = season_means)

fit_seg_naive <- segmented(fit_lm, seg.Z = ~year, psi = events$year)
naive_se <- fit_seg_naive$psi[, "St.Err"]

cat("Standard errors on the 6 simultaneous breakpoints (years):\n")
print(round(naive_se, 1))
cat("\nMax SE:", round(max(naive_se), 1), "years -",
    "this large means at least one breakpoint is not reliably identified.\n")
```

```{r segmented-selected}
fit_seg <- selgmented(fit_lm, seg.Z = ~year, Kmax = 6, type = "bic", msg = FALSE)

cat("BIC-selected number of breakpoints:", nrow(fit_seg$psi), "\n\n")
fit_seg$psi %>% print()
```

**Finding:** forcing 6 simultaneous breakpoints produces standard errors up
to `r round(max(naive_se), 1)` years — several breakpoints are essentially
unidentified with only ~39 annual observations. Letting BIC choose instead
selects `r nrow(fit_seg$psi)` breakpoints, all estimated tightly (SE well
under a year to ~1 year): around
`r paste(round(fit_seg$psi[, "Est."], 0), collapse = ", ")`. Comparing to
the event table: the first is ambiguous between the two early
equipment-advance events (1992/1996); the second lands close to (but
`r round(fit_seg$psi[2, "Est."], 0) - 2003` years after) the 2003-04
COR/MOI Limits event — a real example of the lag we were watching for; the
third (~`r round(fit_seg$psi[3, "Est."], 0)`) doesn't correspond to any
event in our list at all, and is worth flagging as an unexplained
inflection rather than force-fitting it to the nearest regulation.

```{r segmented-plot}
season_means_fitted <- season_means %>%
  mutate(fitted = predict(fit_seg))

p_segmented <- ggplot(season_means_fitted, aes(year, mean_distance)) +
  geom_vline(data = events, aes(xintercept = year, color = category),
             linetype = "dashed", linewidth = 0.5, alpha = 0.6) +
  geom_point(color = "grey50", size = 1) +
  geom_line(aes(y = fitted), color = COL_PRIMARY, linewidth = 1.1) +
  scale_color_manual(values = c(
    "Equipment advance"      = COL_SECONDARY,
    "Regulatory restriction" = COL_ACCENT
  )) +
  labs(
    title    = "Segmented Regression: BIC-Selected Breakpoints",
    subtitle = "Solid line: piecewise-linear fit. Dashed lines: nominal regulatory events.",
    x = NULL, y = "Mean Driving Distance (yards)", color = NULL
  ) +
  THEME_BASE

p_segmented
```

---

# 5. Data-Driven Breakpoint Detection

Two independent methods, neither given the event years, to check whether
purely data-driven breaks land near the known regulations (a form of
external validation for the segmented-regression result above).

## 5.1 PELT on the differenced series

PELT (`changepoint::cpt.mean`) run on the raw distance *level* series would
flag breaks almost everywhere, since the level series has a strong secular
trend baked in — the "mean" of a trending series is never stable for long.
Running it on the *first-differenced* series (year-over-year gain) instead
detects shifts in the *rate* of gain, which is what we actually care about.
We use `minseglen = 3` to avoid the algorithm chasing single-year noise
into an implausibly large number of breaks — without it, PELT returns 8
breakpoints clustered around the already-known 2000-2003 transition, which
is more a symptom of a choppy few years than 8 distinct regime changes.

```{r pelt}
diffs <- diff(season_means$mean_distance)

cpt_fit <- cpt.mean(diffs, method = "PELT", penalty = "BIC", minseglen = 3)
pelt_years <- season_means$year[cpts(cpt_fit) + 1]

cat("PELT breakpoints (years):", paste(pelt_years, collapse = ", "), "\n")
```

## 5.2 Bai-Perron via strucchange

`strucchange::breakpoints()` fits the regression directly (`mean_distance ~ year`)
and finds the BIC-optimal number and location of slope breaks — the more
natural tool for a trending regression, complementing PELT's differenced-series
approach.

```{r bai-perron}
bp <- breakpoints(mean_distance ~ year, data = season_means, h = 0.15)
bp_years <- season_means$year[bp$breakpoints]

cat("Bai-Perron breakpoints (years):", paste(bp_years, collapse = ", "), "\n")
```

## 5.3 CUSUM structural stability test

A visual complement: does the cumulative sum of recursive residuals stay
within its confidence bounds, or cross them (signaling structural
instability)?

```{r cusum}
cusum_fit <- efp(mean_distance ~ year, data = season_means, type = "OLS-CUSUM")
plot(cusum_fit, main = "OLS-CUSUM Structural Stability Test")
```

**Finding:** PELT (`r paste(pelt_years, collapse = ", ")`) and Bai-Perron
(`r paste(bp_years, collapse = ", ")`) roughly agree on an early-1990s
break and a break in the early-to-mid 2000s, but diverge afterward — PELT
finds nothing past 2003, while Bai-Perron detects a further break around
2012, echoing the unexplained ~2014 inflection from the segmented model in
section 4. Both methods' early breaks land close to the 1992 Oversized
Driver rule and the 2000/2003 Pro V1/COR-MOI window — real agreement
between a hypothesis-driven and a purely data-driven approach. The CUSUM
path crossing its bounds confirms the series isn't structurally stable
over the full window, consistent with both breakpoint methods.

---

# 6. Interrupted Time Series (ITS) Per Event

For each event: fit `distance ~ year_centered * post` on a ±5-year window
(3-year post-window for the 2022 event only — flagged explicitly, since
2025 is the last year of data). Rather than assume the effect starts
exactly at the nominal year, we test the `post` split at lags of 0, 1, and
2 years and keep whichever lag fits best (highest R²) — this is where the
"does the effect take a year or two to show up" question gets a direct,
per-event answer rather than a single overall assumption.

```{r its-function}
fit_its_event <- function(event_year, label, category, pre_years = 5) {
  post_years_available <- DATA_END_YEAR - event_year
  post_years <- min(5, post_years_available)
  truncated  <- post_years < 5

  window <- season_means %>%
    filter(year >= event_year - pre_years, year <= event_year + post_years)

  lag_results <- map_dfr(0:2, function(lag) {
    w <- window %>%
      mutate(year_c = year - event_year, post = as.integer(year >= event_year + lag))
    # Skip a lag if it leaves fewer than 2 points on either side of the split
    if (n_distinct(w$post) < 2 || sum(w$post == 0) < 2 || sum(w$post == 1) < 2) return(tibble())
    fit <- lm(mean_distance ~ year_c * post, data = w)
    tibble(lag = lag, r2 = summary(fit)$r.squared, fit = list(fit))
  })

  best <- lag_results %>% slice_max(r2, n = 1, with_ties = FALSE)
  fit  <- best$fit[[1]]
  co   <- tidy(fit)

  tibble(
    year = event_year, label = label, category = category,
    best_lag = best$lag, truncated_post = truncated, post_years_used = post_years,
    level_change = co$estimate[co$term == "post"],
    level_se     = co$std.error[co$term == "post"],
    slope_change = co$estimate[co$term == "year_c:post"],
    slope_se     = co$std.error[co$term == "year_c:post"]
  )
}

its_results <- pmap_dfr(events, function(year, label, category) {
  fit_its_event(year, label, category)
})

its_results %>% print(width = Inf)
```

```{r its-forest-plot}
its_plot_data <- its_results %>%
  mutate(
    level_sig = abs(level_change / level_se) > qt(0.975, df = 8),
    slope_sig = abs(slope_change / slope_se) > qt(0.975, df = 8),
    label_ord = fct_reorder(paste0(label, " (", year, ")"), year)
  )

p_forest_level <- ggplot(its_plot_data, aes(level_change, label_ord, color = category)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_errorbarh(aes(xmin = level_change - 1.96 * level_se, xmax = level_change + 1.96 * level_se), height = 0.2) +
  geom_point(size = 3) +
  scale_color_manual(values = c("Equipment advance" = COL_SECONDARY, "Regulatory restriction" = COL_ACCENT)) +
  labs(title = "Level Change at Each Event (best-fitting lag)", x = "Yards", y = NULL, color = NULL) +
  THEME_BASE

p_forest_slope <- ggplot(its_plot_data, aes(slope_change, label_ord, color = category)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_errorbarh(aes(xmin = slope_change - 1.96 * slope_se, xmax = slope_change + 1.96 * slope_se), height = 0.2) +
  geom_point(size = 3) +
  scale_color_manual(values = c("Equipment advance" = COL_SECONDARY, "Regulatory restriction" = COL_ACCENT)) +
  labs(title = "Slope Change at Each Event (yds/year, best-fitting lag)", x = "Yards/year", y = NULL, color = NULL) +
  THEME_BASE

p_forest_level
p_forest_slope
```

**Finding:** the clearest, most interpretable result is the 2003-04 COR/MOI
event: a significant *level* jump
(+`r round(its_results$level_change[its_results$year==2003], 1)` yds) followed
by a significant *slope deceleration*
(`r round(its_results$slope_change[its_results$year==2003], 2)` yds/yr) —
distance jumped sharply, then growth cooled off, exactly the pattern you'd
expect from a one-time equipment step-change rather than an ongoing
acceleration. 1996 Titanium Drivers shows the opposite pattern — a
significant slope *increase* with no immediate level jump — consistent with
a technology that took a few years to diffuse through the field rather
than an instant swap. The two most recent "restriction" events (2010 Groove
Rule, 2022 Driver Length Limit) show no statistically significant slope
change — no evidence either one meaningfully slowed distance gains (2022's
estimate is also the least trustworthy of the six, given its
`r its_results$post_years_used[its_results$year==2022]`-year, not 5-year,
post-window).

---

# 7. Counterfactual Projection

Project the pre-existing trend forward and compare to what actually
happened — the gap is the "excess" distance gained beyond where the sport
was already headed. The choice of baseline window matters a lot here:
using data through 2002 (the documented "pre-2003" cutoff) already
partially includes the early part of the 2000-2003 acceleration itself, so
we also show a stricter pre-2000 ("before Pro V1") baseline as a robustness
check.

```{r counterfactual}
fit_counterfactual <- function(cutoff_year) {
  pre_data <- season_means %>% filter(year < cutoff_year)
  fit <- lm(mean_distance ~ year, data = pre_data)
  pred <- predict(fit, newdata = season_means, interval = "prediction", level = 0.95)
  bind_cols(season_means, as_tibble(pred)) %>% mutate(baseline = paste0("Pre-", cutoff_year))
}

counterfactuals <- bind_rows(
  fit_counterfactual(2000),
  fit_counterfactual(2003)
)

p_counterfactual <- ggplot(counterfactuals, aes(year)) +
  geom_ribbon(aes(ymin = lwr, ymax = upr, fill = baseline), alpha = 0.15) +
  geom_line(aes(y = fit, color = baseline), linetype = "dashed", linewidth = 0.9) +
  geom_line(aes(y = mean_distance), color = "black", linewidth = 1) +
  scale_color_manual(values = c("Pre-2000" = COL_ACCENT, "Pre-2003" = COL_SECONDARY)) +
  scale_fill_manual(values = c("Pre-2000" = COL_ACCENT, "Pre-2003" = COL_SECONDARY)) +
  labs(
    title    = "Actual Distance vs. Counterfactual Projections",
    subtitle = "Black: actual. Dashed: pre-trend projected forward, with 95% prediction band.",
    x = NULL, y = "Mean Driving Distance (yards)", color = NULL, fill = NULL
  ) +
  THEME_BASE

p_counterfactual
```

```{r counterfactual-gap}
gap_2000 <- counterfactuals %>% filter(baseline == "Pre-2000", year == DATA_END_YEAR) %>%
  mutate(gap = mean_distance - fit) %>% pull(gap)
gap_2003 <- counterfactuals %>% filter(baseline == "Pre-2003", year == DATA_END_YEAR) %>%
  mutate(gap = mean_distance - fit) %>% pull(gap)

slope_pre2000 <- coef(lm(mean_distance ~ year, data = season_means %>% filter(year < 2000)))[["year"]]
slope_pre2003 <- coef(lm(mean_distance ~ year, data = season_means %>% filter(year < 2003)))[["year"]]
```

**Finding:** the two baselines tell different-looking stories, and the
difference is itself the finding. The pre-2000 trend (slope
`r round(slope_pre2000, 2)` yds/yr) projects a `r round(gap_2000, 1)`-yard
gap by `r DATA_END_YEAR` — a real, sizable "excess" beyond where the sport
was headed before Pro V1. The pre-2003 trend (slope
`r round(slope_pre2003, 2)` yds/yr) — already steeper, because it
partially includes the 2000-2002 ramp-up — projects a gap of only
`r round(gap_2003, 1)` yards, since it's extrapolating a rate that already
absorbed part of the acceleration. Neither number is "the" right answer;
together they show the counterfactual gap is highly sensitive to exactly
when you consider the "old normal" to have ended, which is worth keeping in
mind before quoting a single "excess yards" figure anywhere else.

---

# 8. Accuracy — Lighter Parallel Check

The question covers both distance and accuracy. Rather than repeat every
analysis above, we run the same two structural-break tools (segmented
regression, PELT) on `mean_accuracy` alone and compare.

```{r accuracy-check}
fit_lm_acc <- lm(mean_accuracy ~ year, data = season_means)
fit_seg_acc <- selgmented(fit_lm_acc, seg.Z = ~year, Kmax = 4, type = "bic", msg = FALSE)

cat("Accuracy breakpoints (BIC-selected):\n")
print(fit_seg_acc$psi)

diffs_acc <- diff(season_means$mean_accuracy)
cpt_acc <- cpt.mean(diffs_acc, method = "PELT", penalty = "BIC", minseglen = 3)
pelt_acc_years <- season_means$year[cpts(cpt_acc) + 1]
cat("\nAccuracy PELT breakpoints:", if (length(pelt_acc_years) == 0) "none detected" else paste(pelt_acc_years, collapse = ", "), "\n")

acc_1987 <- season_means %>% filter(year == min(year)) %>% pull(mean_accuracy)
acc_peak_year <- fit_seg_acc$psi[1, "Est."]
acc_2025 <- season_means %>% filter(year == max(year)) %>% pull(mean_accuracy)
```

**Finding:** accuracy shows a genuinely different pattern from distance,
not just a mirror image of it. It's non-monotonic — rising from
`r round(acc_1987, 1)`% in 1987 to a peak around
`r round(acc_peak_year, 0)`, then declining substantially before leveling
off (`r round(acc_2025, 1)`% by 2025) — and its own segmented breakpoints
(~`r round(fit_seg_acc$psi[1,"Est."],0)`, ~`r round(fit_seg_acc$psi[2,"Est."],0)`)
don't line up cleanly with distance's. Most tellingly, PELT finds
**`r if (length(pelt_acc_years)==0) "no" else length(pelt_acc_years)`**
rate-of-change breakpoints in accuracy's year-over-year series (vs. 3 for
distance) — accuracy's rate of change has been comparatively gradual and
smooth, with no sharp regime shifts, while distance's has clear, detectable
jumps. That asymmetry itself supports the idea that the technology changes
studied here acted much more directly on distance than on accuracy.

---

# 9. Summary

- **Segmented regression and two independent data-driven methods broadly
  agree**: there's a detectable break in the early 1990s and another
  around the 2000-2003 Pro V1/COR-MOI window — but a further, unexplained
  break around 2012-2014 shows up in two of the three methods that doesn't
  correspond to any documented regulatory event.
- **Effects are not uniformly instant.** The COR/MOI break in the
  segmented model landed a couple of years after the nominal date, and the
  1996 Titanium Drivers event shows up as a slope increase with no
  immediate level jump — consistent with gradual equipment adoption rather
  than an instant switch.
- **The "restriction vs. advance" distinction matters.** The clearest,
  most statistically confident event — COR/MOI Limits — actually shows a
  significant level jump *followed by* a slope slowdown, i.e. exactly the
  kind of one-time step-change-then-plateau you'd expect from a rule that
  capped what had just been unlocked. The two most recent restrictions
  (Groove Rule, Driver Length Limit) show no significant slope effect at
  all — no evidence the rules meaningfully slowed the trend, echoing this
  project's overarching narrative that regulatory interventions have had
  limited effect.
- **The counterfactual "excess yards" number is baseline-dependent** — a
  pre-2000 baseline implies real excess distance gain; a pre-2003 baseline
  implies almost none, because it already absorbs part of the run-up. Be
  explicit about which baseline is being used whenever this number is
  quoted elsewhere (e.g. in a blog post).
- **Accuracy moves differently from distance**, not in lockstep — it rose
  through the mid-1990s, then declined and flattened, with no sharp
  data-driven breakpoints at all. Distance's gains look like a
  technology/rule story; accuracy's decline looks more gradual and
  diffuse.

---

# 10. Session Info

```{r session-info}
sessionInfo()
```
