---
title: "03 — The Distance Premium"
subtitle: "PGA Tour Distance & Speed Analytics"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    df_print: paged
---

# Overview

**Question:** Has the relationship between driving distance and
competitive success (SG_total) strengthened over time?

**Data:** `df_main` (2004–2025, strokes-gained era).

**A preview of what's coming, so the individual sections make sense in
context:** the simplest year-by-year metrics here (Spearman correlation,
annual regression slope) are noisy and non-monotonic — they don't cleanly
answer the question on their own. A pooled mixed-effects model (section 6)
resolves it: yes, the premium has strengthened, and by a lot (roughly
tripled over the window). We show the noisy simple metrics anyway, because
part of the noise itself turns out to be explained by something real (how
"bunched up" the field's distances were in a given year — section 2), and
because the more granular distance-quintile comparison (section 5) gives
an independent, cleaner confirmation of the same trend.

---

# 1. Setup

```{r setup, message=FALSE, warning=FALSE}
# Core
library(tidyverse)
library(here)

# Modeling
library(lme4)         # mixed-effects interaction model (section 6)

# Visualization
library(patchwork)    # scatter triptych (section 8)
library(scales)

# Utilities
library(broom)
library(broom.mixed)  # tidy() for lmer objects

source(here::here("config.R"))

set.seed(2025)

knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 10,
  fig.height = 6
)
```

---

# 2. Load Data

```{r load-data}
df_main <- readRDS(here(PATH_PROCESSED, "df_main.rds"))

cat("Rows:          ", nrow(df_main), "\n")
cat("Unique players:", n_distinct(df_main$player_name), "\n")
cat("Year range:    ", min(df_main$year), "-", max(df_main$year), "\n")
```

## 2.1 Two quick data checks before analyzing

None of the columns used in this notebook have missing values, and
`sg_total` should equal the sum of its four components — worth confirming
rather than assuming, since a real discrepancy would undermine section 4's
component-importance analysis.

```{r data-checks}
key_cols <- c("sg_total", "distance", "dist_vs_field", "dist_quintile",
              "sg_ott", "sg_apr", "sg_arg", "sg_putt", "age")
na_counts <- colSums(is.na(df_main[key_cols]))
cat("NA counts in key columns:\n")
print(na_counts[na_counts > 0])
if (all(na_counts == 0)) cat("(none - all zero)\n")

sg_diff <- df_main$sg_total - (df_main$sg_ott + df_main$sg_apr + df_main$sg_arg + df_main$sg_putt)
cat("\nsg_total vs. sum of 4 components - max abs difference:", round(max(abs(sg_diff)), 3), "\n")
cat("Rows differing by more than 0.2:", sum(abs(sg_diff) > 0.2), "of", nrow(df_main), "\n")
```

**Finding:** zero missing values in any column used here. `sg_total` is
essentially, but not exactly, the sum of its four components (max
discrepancy `r round(max(abs(sg_diff)), 3)`, and only
`r sum(abs(sg_diff) > 0.2)` of `r nrow(df_main)` rows differ by more than
0.2) — clearly just rounding in the source data, not a real data-quality
problem. Safe to treat `sg_total` as the sum of the four components for
section 4's variance-decomposition purposes.

---

# 3. Year-by-Year Correlation: Distance vs. SG_total

The most direct test of "does hitting it farther associate with playing
better" in a given season. We use Spearman (rank) correlation as the
headline metric since it doesn't assume a linear relationship — but before
treating it as the primary metric, it's worth checking whether using ranks
instead of actual yardage changes the picture, particularly in years where
the field's distances might be more tightly bunched together.

```{r yearly-correlation}
yearly_corr <- df_main %>%
  group_by(year) %>%
  summarise(
    spearman_rho     = cor(distance, sg_total, method = "spearman", use = "complete.obs"),
    pearson_raw_dist = cor(distance, sg_total, method = "pearson", use = "complete.obs"),
    pearson_relative = cor(dist_vs_field, sg_total, method = "pearson", use = "complete.obs"),
    sd_distance      = sd(distance, na.rm = TRUE),
    .groups = "drop"
  )

yearly_corr %>% print(n = 10)
```

## 3.1 Does rank vs. magnitude change the story?

```{r rank-vs-magnitude-check}
cat("Pearson (raw distance) vs. Pearson (relative distance) - should be identical:\n")
cat("Max difference:", max(abs(yearly_corr$pearson_raw_dist - yearly_corr$pearson_relative)), "\n\n")

cat("Correlation between Spearman and Pearson series across years:\n")
cat(round(cor(yearly_corr$spearman_rho, yearly_corr$pearson_raw_dist), 3), "\n\n")

cat("Correlation between a year's distance SPREAD and that year's Spearman rho:\n")
cat(round(cor(yearly_corr$sd_distance, yearly_corr$spearman_rho), 3), "\n")
```

```{r yearly-correlation-plot}
p_yearly_corr <- ggplot(yearly_corr, aes(year, spearman_rho)) +
  geom_line(color = COL_PRIMARY, linewidth = 1) +
  geom_point(color = COL_PRIMARY) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  labs(
    title    = "Distance-SG_total Correlation by Year",
    subtitle = "Spearman rank correlation, distance vs. season strokes-gained total",
    x = NULL, y = "Spearman's rho"
  ) +
  THEME_BASE

p_yearly_corr
```

**Finding:** this series is noisy and non-monotonic, not a clean rising
trend — from `r round(yearly_corr$spearman_rho[1], 2)` in
`r yearly_corr$year[1]` up to a peak of
`r round(max(yearly_corr$spearman_rho), 2)` in
`r yearly_corr$year[which.max(yearly_corr$spearman_rho)]`, then back down to
`r round(yearly_corr$spearman_rho[yearly_corr$year == 2024], 2)` in 2024. Two
things worth separating out: (1) using raw distance or field-relative
distance gives *mathematically identical* Pearson correlations within each
year (relative distance is just raw distance shifted by a constant, and
Pearson correlation is invariant to shifts) — confirmed above, max
difference is exactly 0. (2) Spearman and Pearson are 0.98 correlated
across years, so rank-vs-magnitude isn't why the series is noisy. But a
year's distance *spread* (SD) correlates 0.42 with that year's rho — years
where the field's distances are more compressed genuinely do show a weaker
relationship, regardless of how it's measured. That's a real, partial
explanation for the noise here, and another reason to lean on the pooled
model in section 6 rather than any single year's number.

---

# 4. Annual R²: SG_total ~ SG_OTT

How much of the variation in total performance does off-the-tee play (the
component distance most directly feeds) explain, and has that grown?

```{r annual-r2-ott}
r2_by_year <- df_main %>%
  group_by(year) %>%
  summarise(r2 = summary(lm(sg_total ~ sg_ott))$r.squared, .groups = "drop")

p_r2 <- ggplot(r2_by_year, aes(year, r2)) +
  geom_line(color = COL_SECONDARY, linewidth = 1) +
  geom_point(color = COL_SECONDARY) +
  labs(
    title    = "How Much Does Off-the-Tee Play Explain Total Performance?",
    subtitle = "R² from SG_total ~ SG_OTT, fit separately each year",
    x = NULL, y = expression(R^2)
  ) +
  THEME_BASE

p_r2
```

**Finding:** R² ranges from
`r round(min(r2_by_year$r2), 2)` (`r r2_by_year$year[which.min(r2_by_year$r2)]`)
to `r round(max(r2_by_year$r2), 2)` (`r r2_by_year$year[which.max(r2_by_year$r2)]`),
noisy like section 3 but with a visible tilt: the pre-2012 years mostly sit
below 0.3, while 2016 onward mostly sit above 0.3 — off-the-tee play
explains a somewhat larger share of who plays well in the more recent
strokes-gained era than it did at the start of it.

---

# 5. SG Component Importance (Standardized Betas)

Since `sg_total` is essentially the sum of its four components (section
2.1), regressing it on all four **will** produce R² close to 1 by
construction — that's expected, not a modeling success. The informative
part is the *relative size* of the four standardized betas: which
component of the game actually drives the variation in total score, and
has that mix shifted toward off-the-tee (where distance lives)?

```{r component-importance}
beta_by_year <- df_main %>%
  group_by(year) %>%
  mutate(across(c(sg_ott, sg_apr, sg_arg, sg_putt), ~as.numeric(scale(.x)), .names = "z_{.col}")) %>%
  group_modify(~{
    fit <- lm(sg_total ~ z_sg_ott + z_sg_apr + z_sg_arg + z_sg_putt, data = .x)
    tibble(term = names(coef(fit))[-1], beta = coef(fit)[-1], r2 = summary(fit)$r.squared)
  }) %>%
  ungroup() %>%
  mutate(component = str_remove(term, "^z_sg_"))

cat("R2 range across years (expected to be ~1):\n")
print(range(beta_by_year$r2))
```

```{r component-importance-plot}
component_labels <- c(ott = "Off-the-Tee", apr = "Approach", arg = "Around-the-Green", putt = "Putting")

p_components <- ggplot(beta_by_year, aes(year, beta, color = component)) +
  geom_line(linewidth = 1) +
  geom_point() +
  scale_color_manual(values = c(
    ott = COL_PRIMARY, apr = COL_SECONDARY, arg = COL_NEUTRAL, putt = COL_ACCENT
  ), labels = component_labels) +
  labs(
    title    = "Standardized Contribution of Each SG Component to SG_total",
    subtitle = "Larger beta = that part of the game explains more of the spread in total performance",
    x = NULL, y = "Standardized beta", color = NULL
  ) +
  THEME_BASE

p_components
```

```{r component-findings}
ott_2004 <- beta_by_year %>% filter(year == min(year), component == "ott") %>% pull(beta)
ott_2025 <- beta_by_year %>% filter(year == max(year), component == "ott") %>% pull(beta)
apr_2004 <- beta_by_year %>% filter(year == min(year), component == "apr") %>% pull(beta)
apr_2025 <- beta_by_year %>% filter(year == max(year), component == "apr") %>% pull(beta)
top_2025 <- beta_by_year %>% filter(year == max(year)) %>% slice_max(beta, n = 1) %>% pull(component)
```

**Finding:** R² is indeed ~1 in every year (as expected — this confirms the
components decompose the total cleanly, it's not a meaningful model-fit
result). Approach play (`apr`) has the largest standardized beta in both
2004 (`r round(apr_2004, 2)`) and 2025 (`r round(apr_2025, 2)`) — it's
still the single biggest driver of who separates from the pack on
`sg_total`, contrary to what a pure "distance premium" narrative alone
might suggest. But off-the-tee's beta rose from
`r round(ott_2004, 2)` to `r round(ott_2025, 2)`, the largest increase of
any component, narrowing the gap with approach considerably. The game
hasn't become "all about the driver," but the driver's relative importance
has clearly grown.

---

# 6. Distance Quintile SG_total Trajectories

Using the field-relative distance quintiles already engineered in
`00_data_engineering.Rmd` (`dist_quintile`, 1 = shortest, 5 = longest) — a
grouped, less noisy view of the same relationship as sections 3/7.

```{r quintile-trajectories}
quintile_traj <- df_main %>%
  group_by(year, dist_quintile) %>%
  summarise(mean_sg_total = mean(sg_total, na.rm = TRUE), .groups = "drop")

p_quintiles <- ggplot(quintile_traj, aes(year, mean_sg_total, color = factor(dist_quintile))) +
  geom_line(linewidth = 1) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  scale_color_manual(values = COL_QUINTILES, name = "Distance\nQuintile") +
  labs(
    title    = "Mean SG_total by Distance Quintile",
    subtitle = "Quintile 1 = shortest hitters that season, 5 = longest",
    x = NULL, y = "Mean SG_total"
  ) +
  THEME_BASE

p_quintiles
```

```{r quintile-gap}
gap_by_year <- quintile_traj %>%
  filter(dist_quintile %in% c(1, 5)) %>%
  pivot_wider(names_from = dist_quintile, values_from = mean_sg_total, names_prefix = "q") %>%
  mutate(gap = q5 - q1)

any_reversed <- sum(gap_by_year$gap < 0)
gap_early <- gap_by_year %>% filter(year <= 2013) %>% summarise(m = mean(gap)) %>% pull(m)
gap_late  <- gap_by_year %>% filter(year >= 2014) %>% summarise(m = mean(gap)) %>% pull(m)
```

**Finding:** the middle quintiles (2-4) aren't always perfectly ordered in
a given year, but the top quintile (Q5, longest) beats the bottom quintile
(Q1, shortest) in **every single year** of the dataset
(`r any_reversed` reversals out of `r n_distinct(gap_by_year$year)` years).
The gap has clearly widened: averaging
`r round(gap_early, 2)` SG in 2004-2013 vs.
`r round(gap_late, 2)` in 2014-2025 — a clean, independent confirmation
that the longest hitters' advantage over the shortest has grown, matching
the pooled model result in the next section.

---

# 7. Interaction Model: The Rigorous Test

Sections 3-6 all point the same direction but are either noisy (3, 4) or
descriptive rather than inferential (6). This model pools every
player-season, controls for player-level heterogeneity with a random
intercept and for age, and tests the distance-premium trend directly via a
`dist_vs_field:year` interaction. `dist_vs_field` is rescaled to
`dist_vs_field_10` (per 10 yards) so coefficients are directly
interpretable as "SG per 10 yards above the field," and `year` is centered
at its median so the main `dist_vs_field_10` effect describes the premium
at a typical year in the sample, not a linear extrapolation back to year 0.

```{r interaction-model}
df_model <- df_main %>%
  mutate(
    dist_vs_field_10 = dist_vs_field / 10,
    year_c = year - median(year)
  )

fit_interaction <- lmer(
  sg_total ~ dist_vs_field_10 * year_c + age + (1 | player_name),
  data = df_model
)

tidy(fit_interaction, effects = "fixed") %>% print()
```

```{r interaction-findings}
coefs <- tidy(fit_interaction, effects = "fixed")
premium_center <- coefs$estimate[coefs$term == "dist_vs_field_10"]
premium_trend  <- coefs$estimate[coefs$term == "dist_vs_field_10:year_c"]
premium_trend_t <- coefs$statistic[coefs$term == "dist_vs_field_10:year_c"]

year_span <- (max(df_model$year) - min(df_model$year)) / 2
premium_2004 <- premium_center - premium_trend * year_span
premium_2025 <- premium_center + premium_trend * year_span
```

**Finding:** this is the test that actually resolves the question. The
`dist_vs_field_10:year_c` interaction is
`r round(premium_trend, 4)` (t = `r round(premium_trend_t, 2)`) — small per
year, but real and statistically clear once pooled across ~4,000
player-seasons. At the sample's center year, the premium is
`r round(premium_center, 2)` SG per 10 yards above the field; projecting
the trend to the ends of the window implies it's grown from roughly
`r round(premium_2004, 2)` (2004) to roughly `r round(premium_2025, 2)`
(2025) — **the distance premium has roughly
`r round(premium_2025 / premium_2004, 1)`x'd** over the strokes-gained era.
This is the reliable version of what sections 3 and 6 hinted at noisily.

---

# 8. Distance Premium Metric: Annual Slope Trajectory

A year-by-year complement to section 7's single pooled estimate — expect
this to be noisy for the same reasons as section 3 (same underlying
data), with section 7 as the trustworthy summary.

```{r annual-premium}
premium_by_year <- df_main %>%
  group_by(year) %>%
  summarise(slope_per_10yd = coef(lm(sg_total ~ dist_vs_field))[["dist_vs_field"]] * 10, .groups = "drop")

p_premium <- ggplot(premium_by_year, aes(year, slope_per_10yd)) +
  geom_line(color = COL_ACCENT, linewidth = 1) +
  geom_point(color = COL_ACCENT) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  labs(
    title    = "Distance Premium by Year",
    subtitle = "SG_total per 10 yards above the field average, estimated separately each year",
    x = NULL, y = "SG_total per 10 yards"
  ) +
  THEME_BASE

p_premium
```

**Finding:** noisy year-to-year (ranging
`r round(min(premium_by_year$slope_per_10yd), 2)` to
`r round(max(premium_by_year$slope_per_10yd), 2)`), but with the same
visible tilt as section 4's R² series — pre-2012 years cluster lower,
2014 onward cluster higher — corroborating section 7's pooled estimate
without needing to trust any single year's number.

---

# 9. Scatter Triptych: Three Eras

```{r scatter-triptych}
triptych_years <- c(2005, 2013, 2025)

triptych_plots <- map(triptych_years, function(yr) {
  df_main %>%
    filter(year == yr) %>%
    ggplot(aes(distance, sg_total)) +
    geom_point(alpha = 0.5, color = COL_PRIMARY) +
    geom_smooth(method = "lm", se = FALSE, color = COL_ACCENT, linewidth = 0.8) +
    labs(title = as.character(yr), x = "Distance (yds)", y = "SG_total") +
    THEME_BASE
})

wrap_plots(triptych_plots, nrow = 1) +
  plot_annotation(title = "Distance vs. SG_total: Three Eras")
```

**Finding:** the fitted line's slope visibly steepens from 2005 to 2025 — a
direct visual echo of the interaction model's finding in section 7.

---

# 10. Summary

- **The distance premium has strengthened substantially** — the pooled
  mixed-effects model (section 7) is the reliable evidence: the premium has
  gone from roughly 0.1 to roughly 0.3 SG per 10 yards above the field, a
  multi-fold increase, with a statistically clear interaction term behind
  it.
- **The simple year-by-year metrics (sections 3, 8) are too noisy to
  answer this alone** — and part of that noise has a real cause, not just
  small-sample jitter: years with a more compressed field of driving
  distances show weaker year-specific relationships regardless of whether
  measured by rank or magnitude (section 3.1).
- **Approach play, not off-the-tee, is still the single largest
  standardized driver of total performance** (section 4) — but
  off-the-tee's relative importance has grown the most of the four
  components, consistent with (not contradicting) the strengthening
  premium.
- **The longest-hitter quintile has out-performed the shortest-hitter
  quintile every single year** (section 5), with the gap roughly doubling
  from the 2004-2013 average to the 2014-2025 average — an independent,
  low-noise confirmation of the same trend the interaction model finds.

---

# 11. Session Info

```{r session-info}
sessionInfo()
```
