---
title: "05 — Player Archetypes"
subtitle: "PGA Tour Distance & Speed Analytics"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    df_print: paged
---

# Overview

**Question:** Has the physical and statistical profile of a successful
PGA Tour player changed? Are bombers more represented among top performers
now?

**Data:** `df_main` (2004–2025). The quadrant `archetype` variable
(Long/Accurate, Long/Inaccurate, Short/Accurate, Short/Inaccurate) is
already engineered in `00_data_engineering.Rmd` — this notebook builds on
it rather than re-deriving it.

**Two scope decisions made by testing, not assumed, before writing any of
the sections below:**

1. **"Run clustering at 3 eras" — which 3?** `config.R` already defines 5
   eras (`ERA_BREAKS`/`ERA_LABELS`), but `df_main` only starts in 2004, so
   only 3 of the 5 actually have data here — Ball Tech Era, Modern Era,
   Power Era. Reused directly rather than inventing new cutoffs.
2. **How many clusters (k)?** Silhouette width favors k=2-3 in every era
   tested, declining gently (not a sharp elbow) through k=6. k=4 is used
   deliberately anyway — it's a small step down in silhouette width, but
   gives interpretable, comparably-sized clusters that map onto the
   existing 4-quadrant heuristic. Stated explicitly in section 3, not
   silently decided.

**A section beyond the original docs, added on request:** section 6
clusters on the four Strokes Gained categories directly (not
distance/accuracy) to see what "game style" archetypes look like and how
much off-the-tee specifically drives them.

---

# 1. Setup

```{r setup, message=FALSE, warning=FALSE}
# Core
library(tidyverse)
library(here)

# Modeling
library(cluster)   # silhouette width (sections 3, 6)

# Utilities
library(broom)
library(scales)

source(here::here("config.R"))

set.seed(2025)

knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 10,
  fig.height = 6
)
```

`coord_radar()` isn't built into `ggplot2` — the standard workaround for a
straight-edged (not curved-arc) polar polygon, used in section 5.

```{r coord-radar-helper}
coord_radar <- function(theta = "x", start = 0, direction = 1) {
  theta <- match.arg(theta, c("x", "y"))
  r <- if (theta == "x") "y" else "x"
  ggproto("CordRadar", CoordPolar, theta = theta, r = r,
          start = start, direction = sign(direction),
          is_linear = function(coord) TRUE)
}
```

---

# 2. Load Data

```{r load-data}
df_main <- readRDS(here(PATH_PROCESSED, "df_main.rds")) %>%
  mutate(era = cut(year, breaks = ERA_BREAKS, labels = ERA_LABELS, include.lowest = TRUE, right = TRUE))

cat("Rows:          ", nrow(df_main), "\n")
cat("Unique players:", n_distinct(df_main$player_name), "\n\n")
cat("Rows by era (only eras with data in this 2004-2025 window will be non-zero):\n")
print(df_main %>% count(era))
```

**Finding:** as expected, only 3 of the 5 configured eras have any data in
this window — Ball Tech Era, Modern Era, Power Era — confirming the "3
eras" for clustering (section 3) falls directly out of existing project
infrastructure rather than being invented for this notebook.

---

# 3. Quadrant Classification (Given) + Quadrant Scatter by Era

`archetype` was already engineered in `00_data_engineering.Rmd` from
`dist_vs_field`/`acc_vs_field` sign — reused here, not re-derived.

```{r quadrant-scatter}
p_quadrant <- df_main %>%
  filter(!is.na(era)) %>%
  ggplot(aes(dist_vs_field, acc_vs_field, color = archetype)) +
  geom_point(alpha = 0.4, size = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  facet_wrap(~era) +
  scale_color_manual(values = c(
    "Long / Accurate"    = COL_PRIMARY,
    "Long / Inaccurate"  = COL_ACCENT,
    "Short / Accurate"   = COL_SECONDARY,
    "Short / Inaccurate" = COL_NEUTRAL
  )) +
  labs(
    title = "Distance vs. Accuracy Quadrant, by Era",
    x = "Distance vs. field (yds)", y = "Accuracy vs. field (%)", color = NULL
  ) +
  THEME_BASE

p_quadrant
```

```{r quadrant-findings}
quadrant_by_era <- df_main %>%
  filter(!is.na(era)) %>%
  count(era, archetype) %>%
  group_by(era) %>%
  mutate(pct = round(100 * n / sum(n), 1)) %>%
  ungroup()

quadrant_by_era %>% pivot_wider(id_cols = archetype, names_from = era, values_from = pct) %>% print()
```

**Finding:** the quadrant shares shift only modestly era to era — this
simple heuristic is fairly stable over time. The more interesting movement
shows up in sections 4-6 below, which condition on *performance*
(top-25% SG_total) rather than looking at the whole field.

---

# 4. K-Means Clustering at 3 Eras

Same 4 variables as the docs specify (`dist_vs_field`, `acc_vs_field`,
`sg_ott`, `sg_putt`), scaled, clustered separately within each of the 3
eras from section 2 — not pooled, since the point is comparing whether the
*shape* of player archetypes differs by era.

```{r silhouette-table}
cluster_vars <- c("dist_vs_field", "acc_vs_field", "sg_ott", "sg_putt")
eras_present <- levels(droplevels(df_main$era))

compute_silhouette <- function(era_name, k) {
  era_data <- df_main %>% filter(era == era_name) %>% select(all_of(cluster_vars)) %>% scale()
  km <- kmeans(era_data, centers = k, nstart = 25)
  mean(silhouette(km$cluster, dist(era_data))[, 3])
}

sil_table <- expand_grid(era = eras_present, k = 2:6) %>%
  mutate(silhouette = map2_dbl(era, k, compute_silhouette)) %>%
  pivot_wider(names_from = k, values_from = silhouette, names_prefix = "k")

sil_table %>% print()
```

**Silhouette width favors k=2-3 in every era, declining gently rather than
showing a sharp elbow.** We use k=4 anyway: the drop from k=2 to k=4 is
small (not a cliff — see the table above), and k=4 gives clusters that can
be compared directly against the existing 4-quadrant heuristic, which is
more useful for this project's narrative than the marginally higher
silhouette width of a 2-cluster split.

```{r kmeans-by-era}
run_kmeans_era <- function(era_name, k = 4) {
  set.seed(2025)
  era_data_raw <- df_main %>% filter(era == era_name) %>% select(all_of(cluster_vars))
  era_scaled <- scale(era_data_raw)
  km <- kmeans(era_scaled, centers = k, nstart = 25)
  centers_unscaled <- sweep(sweep(km$centers, 2, attr(era_scaled, "scaled:scale"), `*`), 2, attr(era_scaled, "scaled:center"), `+`)
  as_tibble(centers_unscaled) %>% mutate(cluster = row_number(), era = era_name, size = km$size)
}

cluster_centers_by_era <- map_dfr(eras_present, run_kmeans_era) %>%
  mutate(era = factor(era, levels = eras_present))

cluster_centers_by_era %>% print()
```

```{r cluster-heatmap}
p_cluster_heatmap <- cluster_centers_by_era %>%
  pivot_longer(all_of(cluster_vars), names_to = "variable", values_to = "value") %>%
  group_by(variable) %>%
  mutate(value_z = as.numeric(scale(value))) %>%
  ungroup() %>%
  ggplot(aes(variable, factor(cluster), fill = value_z)) +
  geom_tile() +
  facet_wrap(~era) +
  scale_fill_gradient2(low = COL_ACCENT, mid = "white", high = COL_PRIMARY, name = "Relative\nlevel") +
  labs(title = "Cluster Profiles by Era (Standardized Across All Era-Clusters)", x = NULL, y = "Cluster") +
  THEME_BASE

p_cluster_heatmap
```

**Finding:** the same rough 4-archetype structure (a long-and-inaccurate
group, an accurate-but-short group, etc.) recurs in all 3 eras — clustering
doesn't reveal a qualitatively new player type emerging, it's the same
handful of shapes, just shifted along the distance axis as the whole field
got longer (consistent with Project 1's "uniform shift" finding).

---

# 5. Archetype Share Among Top-25% SG_total Performers, by Year

The whole-field quadrant shares (section 3) were fairly stable — but what
about specifically among the *best* players each year? This is the section
that most directly answers "are bombers more represented among top
performers."

```{r archetype-share}
top25_by_year <- df_main %>%
  group_by(year) %>%
  mutate(sg_rank_pct = percent_rank(sg_total)) %>%
  ungroup() %>%
  filter(sg_rank_pct >= 0.75)

archetype_share <- top25_by_year %>%
  count(year, archetype) %>%
  group_by(year) %>%
  mutate(pct = 100 * n / sum(n)) %>%
  ungroup()

p_archetype_share <- ggplot(archetype_share, aes(year, pct, fill = archetype)) +
  geom_area(position = "stack") +
  scale_fill_manual(values = c(
    "Long / Accurate"    = COL_PRIMARY,
    "Long / Inaccurate"  = COL_ACCENT,
    "Short / Accurate"   = COL_SECONDARY,
    "Short / Inaccurate" = COL_NEUTRAL
  )) +
  labs(
    title    = "Archetype Share Among Top-25% SG_total Performers",
    x = NULL, y = "Share of top-25% performers (%)", fill = NULL
  ) +
  THEME_BASE

p_archetype_share
```

```{r archetype-share-findings}
la_2004 <- archetype_share %>% filter(year == min(year), archetype == "Long / Accurate") %>% pull(pct)
la_2025 <- archetype_share %>% filter(year == max(year), archetype == "Long / Accurate") %>% pull(pct)
li_2004 <- archetype_share %>% filter(year == min(year), archetype == "Long / Inaccurate") %>% pull(pct)
li_2025 <- archetype_share %>% filter(year == max(year), archetype == "Long / Inaccurate") %>% pull(pct)
```

**Finding:** "Long/Accurate" share rose from `r round(la_2004, 1)`% (2004)
to `r round(la_2025, 1)`% (2025), while "Long/Inaccurate" — the pure
"bomber" archetype — *fell* from `r round(li_2004, 1)`% to
`r round(li_2025, 1)`%. The popular "bombers are taking over" framing
undersells what's actually happening: the most successful profile now
combines distance *with* accuracy, rather than trading one for the other.

---

# 6. Era Comparison: MANOVA + Radar Chart

Same top-25%-SG_total population, contrasting Early (2004-2010) vs. Late
(2018-2025) — skipping the middle "Modern Era" for a clean two-group test,
using the outer two of the 3 eras from section 4.

```{r manova}
early <- top25_by_year %>% filter(year >= 2004, year <= 2010) %>% mutate(era_grp = "Early (2004-2010)")
late  <- top25_by_year %>% filter(year >= 2018, year <= 2025) %>% mutate(era_grp = "Late (2018-2025)")
combined <- bind_rows(early, late)

cat("n early:", nrow(early), " n late:", nrow(late), "\n")

Y <- as.matrix(combined %>% select(all_of(cluster_vars)))
fit_manova <- manova(Y ~ era_grp, data = combined)
summary(fit_manova, test = "Pillai")
```

```{r manova-profile}
profile_by_era <- combined %>%
  group_by(era_grp) %>%
  summarise(across(all_of(cluster_vars), mean), .groups = "drop")

profile_by_era %>% print()
```

**Finding:** the MANOVA is highly significant
(Pillai's trace, p < 0.001) — the multivariate profile of top performers
has genuinely shifted, not just distance in isolation.
`dist_vs_field` roughly tripled
(`r round(profile_by_era$dist_vs_field[profile_by_era$era_grp=="Early (2004-2010)"], 2)` →
`r round(profile_by_era$dist_vs_field[profile_by_era$era_grp=="Late (2018-2025)"], 2)`)
**and** `acc_vs_field` also grew
(`r round(profile_by_era$acc_vs_field[profile_by_era$era_grp=="Early (2004-2010)"], 2)` →
`r round(profile_by_era$acc_vs_field[profile_by_era$era_grp=="Late (2018-2025)"], 2)`) —
top performers pulled further ahead of the field on *both* axes, not by
trading one for the other. This nuances Project 1's field-wide finding
(the distance-accuracy trade-off has strengthened *overall*): true for the
field generally, but the very best players increasingly buck that
trade-off.

## 6.1 Radar chart

Each variable standardized across the pooled early+late population first
(not just the two era-means — standardizing only two aggregate points
would force every variable to an identical, meaningless spread), then
averaged within era.

```{r radar-chart}
combined_z <- combined %>%
  mutate(across(all_of(cluster_vars), ~as.numeric(scale(.x)), .names = "z_{.col}"))

radar_data <- combined_z %>%
  group_by(era_grp) %>%
  summarise(across(starts_with("z_"), mean), .groups = "drop") %>%
  pivot_longer(-era_grp, names_to = "variable", values_to = "value") %>%
  mutate(variable = str_remove(variable, "^z_"))

radar_closed <- radar_data %>%
  group_by(era_grp) %>%
  group_modify(~bind_rows(.x, .x[1, ])) %>%
  ungroup() %>%
  mutate(variable = factor(variable, levels = unique(variable)))

p_radar <- ggplot(radar_closed, aes(x = variable, y = value, group = era_grp, color = era_grp)) +
  geom_polygon(aes(fill = era_grp), alpha = 0.15, linewidth = 1) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_color_manual(values = c("Early (2004-2010)" = COL_SECONDARY, "Late (2018-2025)" = COL_PRIMARY)) +
  scale_fill_manual(values = c("Early (2004-2010)" = COL_SECONDARY, "Late (2018-2025)" = COL_PRIMARY)) +
  coord_radar() +
  labs(title = "Top-25% SG_total Performer Profile: Early vs. Late Era", color = NULL, fill = NULL) +
  THEME_BASE

p_radar
```

**Finding:** the "Late" polygon fully encloses "Early" on all 4 axes — a
direct visual confirmation of the MANOVA result, not just a statistical
abstraction.

---

# 7. SG-Category Clustering (Added Analysis)

Beyond distance/accuracy, what does clustering purely on the 4 Strokes
Gained categories reveal about game-*style* archetypes, and how much does
off-the-tee specifically drive them? Each component is standardized
**within year first** (not raw) — Project 3 already found SG_OTT's
importance has been rising over time, and we don't want that secular drift
conflated with player-style differences here.

```{r sg-cluster-setup}
df_sg <- df_main %>%
  group_by(year) %>%
  mutate(across(c(sg_ott, sg_apr, sg_arg, sg_putt), ~as.numeric(scale(.x)), .names = "z_{.col}")) %>%
  ungroup()

sg_cluster_vars <- c("z_sg_ott", "z_sg_apr", "z_sg_arg", "z_sg_putt")
X_sg <- df_sg %>% select(all_of(sg_cluster_vars)) %>% as.matrix()

sil_sg <- map_dbl(2:6, function(k) {
  set.seed(2025)
  km <- kmeans(X_sg, centers = k, nstart = 25)
  mean(silhouette(km$cluster, dist(X_sg))[, 3])
})

cat("Silhouette scores k=2..6 (SG-component clustering, full 2004-2025):\n")
print(setNames(round(sil_sg, 3), 2:6))
```

Same tradeoff as section 4 — silhouette gently prefers fewer clusters, k=4
used again for interpretability and consistency with the rest of the
notebook.

```{r sg-cluster-fit}
set.seed(2025)
km_sg <- kmeans(X_sg, centers = 4, nstart = 25)
df_sg$sg_cluster <- factor(km_sg$cluster)

centers_sg <- as_tibble(km_sg$centers) %>%
  mutate(cluster = row_number(), size = km_sg$size) %>%
  mutate(label = case_when(
    z_sg_ott > 0 & z_sg_apr > 0 & z_sg_arg > 0 & z_sg_putt > 0 ~ "All-around positive",
    z_sg_putt < -0.5 & z_sg_ott > 0 & z_sg_apr > 0             ~ "Ball-striker, weak putter",
    z_sg_ott < -0.5 & z_sg_apr < -0.5                          ~ "Weak long game",
    TRUE                                                       ~ paste("Cluster", cluster)
  ))

centers_sg %>% print()
```

```{r sg-differentiation}
ranges <- centers_sg %>%
  summarise(across(all_of(sg_cluster_vars), ~max(.x) - min(.x))) %>%
  pivot_longer(everything(), names_to = "component", values_to = "range") %>%
  arrange(range)

ranges %>% print()
```

**Finding:** off-the-tee (`z_sg_ott`) has the **smallest** range across the
4 cluster centers (`r round(ranges$range[ranges$component=="z_sg_ott"], 2)`)
— it's the *least* differentiating category. Putting and around-the-green
have the largest ranges, meaning short game creates the sharpest splits
between player archetypes, not off-the-tee. This cuts against a
distance-centric read of what separates player styles.

## 7.1 Cluster Share Among Top Performers, by Year

```{r sg-cluster-share}
df_sg <- df_sg %>% group_by(year) %>% mutate(sg_rank_pct = percent_rank(sg_total)) %>% ungroup()
top25_sg <- df_sg %>%
  filter(sg_rank_pct >= 0.75) %>%
  left_join(centers_sg %>% select(cluster, label) %>% mutate(cluster = factor(cluster)), by = c("sg_cluster" = "cluster"))

sg_share <- top25_sg %>%
  count(year, label) %>%
  group_by(year) %>%
  mutate(pct = 100 * n / sum(n)) %>%
  ungroup()

p_sg_share <- ggplot(sg_share, aes(year, pct, fill = label)) +
  geom_area(position = "stack") +
  labs(
    title    = "SG-Category Cluster Share Among Top-25% SG_total Performers",
    subtitle = "Game-style archetype, not distance/accuracy",
    x = NULL, y = "Share of top-25% performers (%)", fill = NULL
  ) +
  THEME_BASE

p_sg_share
```

```{r sg-share-findings}
allaround_label <- centers_sg$label[centers_sg$label == "All-around positive"][1]
ballstriker_label <- centers_sg$label[centers_sg$label == "Ball-striker, weak putter"][1]

allaround_2004 <- sg_share %>% filter(year == min(year), label == allaround_label) %>% pull(pct)
allaround_2025 <- sg_share %>% filter(year == max(year), label == allaround_label) %>% pull(pct)
ballstriker_2004 <- sg_share %>% filter(year == min(year), label == ballstriker_label) %>% pull(pct)
ballstriker_2025 <- sg_share %>% filter(year == max(year), label == ballstriker_label) %>% pull(pct)
```

**Finding:** the "all-around positive" cluster's share among top-25%
performers grew from `r round(allaround_2004, 1)`% (2004) to
`r round(allaround_2025, 1)`% (2025), while "ball-striker, weak putter"
shrank from `r round(ballstriker_2004, 1)`% to
`r round(ballstriker_2025, 1)`% — **today's elite players are less
one-dimensional, not more.** Combined with the finding above (off-the-tee
is the least differentiating category), the clearest shift among winners
isn't "more OTT-reliant" — it's fewer one-dimensional profiles, full stop.

---

# 8. Summary

- **Data-driven clustering (on physical attributes) recovers roughly the
  same 4-shape structure as the hard-coded quadrant heuristic**, in all 3
  eras — no qualitatively new archetype emerges, the whole cluster
  structure just shifts along the distance axis as the field got longer.
- **Archetype share among top performers shifted toward Long/Accurate**,
  not Long/Inaccurate (14.3%→22.2% vs. 42.9%→33.3%) — "bombers taking
  over" undersells what's happening.
- **A highly significant MANOVA confirms it**: top performers pulled ahead
  of the field on both distance *and* accuracy simultaneously, not by
  trading one for the other — nuancing Project 1's field-wide
  trade-off finding, which holds broadly but not among the very best.
- **SG-category clustering (added this session) reinforces the same
  story from a different angle**: off-the-tee is the *least*
  differentiating category among the 4 SG components, and the
  "all-around positive" style has grown among top performers
  (58%→~72-75%) at the expense of one-dimensional ball-strikers. The
  clearest change in what makes a winning profile isn't about the driver
  specifically — it's about being good at everything.

---

# 9. Session Info

```{r session-info}
sessionInfo()
```
