---
title: "08 — Distance Premium and Floor Mechanisms"
subtitle: "PGA Tour Distance & Speed Analytics"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    df_print: paged
---

# Overview

A follow-up to Projects 3 and 4, built after cross-checking a larger
proposed analysis list against what's already done — several of the
originally proposed pieces (the raw annual premium, the quintile gap, the
tiered floor trends) already exist in Projects 3 and 4 and aren't
rebuilt here. This notebook adds two genuinely new angles:

1. **Is the distance premium a real, direct effect, or partly a proxy for
   overall skill?** (Section 2 — the conditional premium.)
2. **Can short-game skill actually compensate for below-average speed, and
   has that gotten harder as the floor has risen?** (Section 4 — the
   compensation ceiling.)

---

# 1. Setup

```{r setup, message=FALSE, warning=FALSE}
# Core
library(tidyverse)
library(here)

# Utilities
library(broom)
library(scales)

source(here::here("config.R"))

set.seed(2025)

knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 10,
  fig.height = 6
)
```

```{r load-data}
df_main <- readRDS(here(PATH_PROCESSED, "df_main.rds")) %>% filter(player_name != "Patrick Sheehan")
df_speed <- df_main %>% filter(year >= SPEED_ERA_START)

cat("Rows:", nrow(df_main), " Speed-era rows:", nrow(df_speed), "\n")
```

---

# 2. Is the Distance Premium Real, or a Proxy for Overall Skill?

Project 3 found the *raw* distance premium (`sg_total ~ dist_vs_field`)
roughly tripled from 2004 to 2025. But that raw premium doesn't control
for anything else — if today's long hitters are increasingly *also* good
at approach play, scrambling, and putting (which Project 5's MANOVA found
they are), some of the raw premium's apparent growth could reflect
distance becoming a better *proxy* for all-around skill, not distance
itself paying off more on its own. This section tests that directly by
adding the other three SG components as controls and seeing whether the
distance coefficient survives.

```{r conditional-premium}
premium_comparison <- df_main %>%
  group_by(year) %>%
  summarise(
    raw_premium       = coef(lm(sg_total ~ dist_vs_field))[["dist_vs_field"]] * 10,
    conditional_premium = coef(lm(sg_total ~ dist_vs_field + sg_apr + sg_arg + sg_putt))[["dist_vs_field"]] * 10,
    .groups = "drop"
  )

premium_comparison %>% print(n = 10)
```

```{r conditional-premium-plot}
premium_long <- premium_comparison %>%
  pivot_longer(-year, names_to = "type", values_to = "premium")

p_premium_comparison <- ggplot(premium_long, aes(year, premium, color = type)) +
  geom_line(linewidth = 1) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  scale_color_manual(
    values = c(raw_premium = COL_ACCENT, conditional_premium = COL_PRIMARY),
    labels = c(raw_premium = "Raw premium", conditional_premium = "Conditional premium (controls for APR/ARG/Putt)")
  ) +
  labs(
    title    = "Raw vs. Conditional Distance Premium",
    subtitle = "SG_total per 10 yards above the field average",
    x = NULL, y = "SG_total per 10 yards", color = NULL
  ) +
  THEME_BASE

p_premium_comparison
```

```{r conditional-premium-findings}
corr_raw_cond <- cor(premium_comparison$raw_premium, premium_comparison$conditional_premium)
raw_2004 <- premium_comparison$raw_premium[premium_comparison$year == min(premium_comparison$year)]
cond_2004 <- premium_comparison$conditional_premium[premium_comparison$year == min(premium_comparison$year)]
raw_sd <- sd(premium_comparison$raw_premium)
cond_sd <- sd(premium_comparison$conditional_premium)
```

**Finding:** the two series correlate only
`r round(corr_raw_cond, 2)` — they tell meaningfully different stories.
In 2004, the raw premium was
`r round(raw_2004, 2)` but the conditional premium was already
`r round(cond_2004, 2)` — over 3x higher. And the conditional premium is
far more *stable* across the whole window (SD =
`r round(cond_sd, 3)`) than the raw premium (SD =
`r round(raw_sd, 3)`), which visibly trends upward. **This means a
meaningful part of the raw premium's apparent growth reflects distance
becoming a better proxy for overall skill** — today's long hitters are
increasingly also the accurate, good-short-game players (Project 5) — **not
that a yard of distance, skill held constant, has become dramatically more
valuable on its own.** The "pure" distance effect was already substantial
in 2004 and has stayed roughly in the same range since.

## 2.1 Reference table

```{r pull-quote-table}
premium_comparison %>%
  filter(year %in% c(2004, 2010, 2015, 2020, 2025)) %>%
  mutate(across(c(raw_premium, conditional_premium), ~round(.x, 2))) %>%
  print()
```

Project 3's pooled mixed-effects model (§7) put the raw premium at ~0.12
(2004) rising to ~0.30 (2025) — roughly a 2.5x increase. This section adds
the missing context: conditional on other skills, the increase is much
smaller, because a large part of the raw trend is compositional (the kind
of player who hits it long has changed), not purely mechanical (a yard of
distance itself paying off more).

---

# 3. Setting Up the Compensation Question

Can a below-average-speed player still compete at an elite level by being
excellent at the short game? And if so, has that gotten harder as the
competitive floor has risen (Project 4)? Restricting to the speed era
(`SPEED_ERA_START`+, since `ball_speed` isn't reliable before 2007).

```{r bottom-speed-setup}
df_bottom_speed <- df_speed %>%
  group_by(year) %>%
  mutate(speed_rank_pct = percent_rank(ball_speed), sg_rank = rank(-sg_total)) %>%
  ungroup() %>%
  filter(speed_rank_pct <= 0.25)

cat("Bottom-25%-ball-speed player-seasons:", nrow(df_bottom_speed), "\n")
```

## 3.1 Does short game actually help this group?

```{r compensation-regression}
fit_compensation <- lm(sg_total ~ sg_putt + sg_apr, data = df_bottom_speed)
tidy(fit_compensation, conf.int = TRUE) %>% print()
```

**Finding:** among bottom-25%-speed players specifically, both `sg_putt`
and `sg_apr` are strong, highly significant predictors of `sg_total`
(coefficients both well above 0.9, p<0.001) — short-game and approach
skill genuinely do help a below-average-speed player post a good overall
season. The compensation mechanism is real, not wishful thinking.

## 3.2 Has it gotten harder to compensate all the way to elite status?

Real short-game skill helping *some* is different from it being enough to
reach the *top* of the field. This tracks the share of bottom-25%-speed
players each year who still make the top 50 in `sg_total`.

```{r compensation-ceiling}
compensation_ceiling <- df_bottom_speed %>%
  group_by(year) %>%
  summarise(n_slow = n(), pct_top50 = mean(sg_rank <= 50) * 100, .groups = "drop")

p_ceiling <- ggplot(compensation_ceiling, aes(year, pct_top50)) +
  geom_line(color = COL_ACCENT, linewidth = 1) +
  geom_point(color = COL_ACCENT) +
  geom_smooth(method = "lm", se = TRUE, color = COL_PRIMARY, linewidth = 0.7, linetype = "dashed") +
  labs(
    title    = "Compensation Ceiling: Can Slow Hitters Still Reach the Top 50?",
    subtitle = "% of bottom-25%-ball-speed players each year who reach top-50 SG_total",
    x = NULL, y = "% reaching top 50"
  ) +
  THEME_BASE

p_ceiling
```

```{r ceiling-findings}
ceiling_trend <- tidy(lm(pct_top50 ~ year, data = compensation_ceiling)) %>% filter(term == "year")
early_ceiling <- compensation_ceiling %>% filter(year <= 2009) %>% summarise(m = mean(pct_top50)) %>% pull(m)
late_ceiling <- compensation_ceiling %>% filter(year >= 2020) %>% summarise(m = mean(pct_top50)) %>% pull(m)
```

**Finding:** the trend is negative
(`r round(ceiling_trend$estimate, 3)`%/year) and **borderline
significant** (p = `r round(ceiling_trend$p.value, 3)`) — worth stating
plainly rather than rounding up to "significant" or dismissing as noise.
The early-era average (2007-2009: `r round(early_ceiling, 1)`%) is
noticeably higher than the recent-era average (2020-2025:
`r round(late_ceiling, 1)`%). Read together with section 3.1: short-game
compensation still genuinely works, but it's gotten harder to compensate
*all the way* to elite status as the overall floor has risen (Project 4) —
consistent with, not contradicting, everything else this project has
found about the rising floor.

---

# 4. Summary

- **The distance premium's apparent 2.5x growth (Project 3) partly
  reflects distance becoming a better proxy for overall skill**, not a
  dramatic increase in distance's own payoff — the conditional premium
  (controlling for other SG components) was already substantial in 2004
  and has stayed comparatively stable since.
- **Short-game skill can genuinely help a below-average-speed player
  compete** — but the share who parlay that into a top-50 season has
  declined (borderline significant), consistent with a rising floor that's
  outpacing what compensation alone can offset.

---

# 5. Session Info

```{r session-info}
sessionInfo()
```
