---
title: "10 — Key Visuals"
subtitle: "PGA Tour Distance & Speed Analytics"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    df_print: paged
---

# Overview

Clean, publication- and social-ready visuals distilled from Projects 1-9.
Minimal prose — one visual per section with a short plain-language caption
underneath. See `docs/project_context.md` for the full analytical context
behind each chart.

---

# Setup

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(here)
library(patchwork)
library(scales)

source(here::here("config.R"))

df_main <- readRDS(here(PATH_PROCESSED, "df_main.rds")) %>% filter(player_name != "Patrick Sheehan")
df_trends <- readRDS(here(PATH_PROCESSED, "df_trends.rds"))

knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 10,
  fig.height = 7
)
```

---

# Section 1: Distributional Trends (2007-2025)

```{r section1-data}
df_speed <- df_main %>% filter(year >= SPEED_ERA_START)

metric_labels <- c(
  distance = "Driving Distance (yds)",
  carry_distance = "Carry Distance (yds)",
  ball_speed = "Ball Speed (mph)",
  chs = "Clubhead Speed (mph)"
)

percentile_data <- map_dfr(names(metric_labels), function(m) {
  df_speed %>%
    group_by(year) %>%
    summarise(
      p10 = quantile(.data[[m]], 0.10),
      p25 = quantile(.data[[m]], 0.25),
      p50 = quantile(.data[[m]], 0.50),
      p75 = quantile(.data[[m]], 0.75),
      p90 = quantile(.data[[m]], 0.90),
      .groups = "drop"
    ) %>%
    mutate(metric = metric_labels[[m]])
}) %>%
  mutate(metric = factor(metric, levels = metric_labels))
```

```{r section1-plot}
p_distributions <- ggplot(percentile_data, aes(year)) +
  geom_ribbon(aes(ymin = p10, ymax = p90), fill = COL_PRIMARY, alpha = 0.15) +
  geom_ribbon(aes(ymin = p25, ymax = p75), fill = COL_PRIMARY, alpha = 0.3) +
  geom_line(aes(y = p50), color = COL_PRIMARY, linewidth = 1.2) +
  facet_wrap(~metric, scales = "free_y") +
  labs(x = NULL, y = NULL) +
  THEME_BASE

p_distributions
```

*(Range annotation removed for now — sparkline insets were too distracting; revisit how to convey band-width stability after the rest of the notebook is built.)*

**The full distribution of driving power metrics has shifted upward
across all four measures since 2007.** Band width — the gap between the
10th and 90th percentile — has remained roughly stable, indicating a
uniform field-wide shift rather than the elite pulling away from the
rest.

```{r section1-save}
ggsave(here(PATH_FIGURES, "publication", "10_distributional_trends.png"),
       p_distributions, width = 10, height = 7, dpi = 300)
ggsave(here(PATH_FIGURES, "social", "10_distributional_trends.png"),
       p_distributions, width = 10, height = 7, dpi = 150)
```

---

# Section 2: Elite Performer Power Trends

```{r section2-data}
top25_speed <- df_main %>%
  filter(year >= SPEED_ERA_START) %>%
  group_by(year) %>%
  slice_max(sg_total, n = 25) %>%
  summarise(
    mean_ball_speed = mean(ball_speed),
    min_ball_speed  = min(ball_speed),
    max_ball_speed  = max(ball_speed),
    .groups = "drop"
  )

tour_speed <- df_main %>%
  filter(year >= SPEED_ERA_START) %>%
  group_by(year) %>%
  summarise(mean_ball_speed = mean(ball_speed), .groups = "drop")

top25_ott <- df_main %>%
  group_by(year) %>%
  slice_max(sg_total, n = 25) %>%
  summarise(mean_sg_ott = mean(sg_ott), .groups = "drop")
```

```{r section2-plot}
shared_xlim <- range(df_main$year)

p_elite_speed <- ggplot(top25_speed, aes(year, mean_ball_speed)) +
  geom_ribbon(aes(ymin = min_ball_speed, ymax = max_ball_speed), fill = COL_PRIMARY, alpha = 0.2) +
  geom_line(aes(color = "Top 25 (SG Total)"), linewidth = 1) +
  geom_point(color = COL_PRIMARY) +
  geom_line(data = tour_speed, aes(color = "Tour average"), linewidth = 0.8, linetype = "dashed") +
  scale_x_continuous(limits = shared_xlim) +
  scale_color_manual(values = c("Top 25 (SG Total)" = COL_PRIMARY, "Tour average" = COL_NEUTRAL), name = NULL) +
  labs(x = NULL, y = "Ball Speed (mph)") +
  THEME_BASE +
  theme(legend.position = "top")

p_elite_ott <- ggplot(top25_ott, aes(year, mean_sg_ott)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_line(color = COL_SECONDARY, linewidth = 1) +
  geom_point(color = COL_SECONDARY) +
  scale_x_continuous(limits = shared_xlim) +
  scale_y_continuous(limits = c(-0.5, 0.5)) +
  labs(x = NULL, y = "SG: Off the Tee") +
  THEME_BASE

p_elite <- p_elite_speed / p_elite_ott
p_elite
```

**Among the tour's top 25 performers by strokes gained, average ball
speed has risen steadily since 2007** — outpacing the tour-wide average
(dashed line) through the middle years before the field closes the gap
again by 2025. The shaded band shows the range from the slowest to
fastest in this elite group each year; the SG: Off the Tee panel is
fixed to a common ±0.5 scale so the size of the effect can be read in
context, not stretched to fill the frame.

```{r section2-save}
ggsave(here(PATH_FIGURES, "publication", "10_elite_performer_power_trends.png"),
       p_elite, width = 10, height = 7, dpi = 300)
ggsave(here(PATH_FIGURES, "social", "10_elite_performer_power_trends.png"),
       p_elite, width = 10, height = 7, dpi = 150)
```

---

# Section 3: Age Composition of Top 100 Performers

```{r section3-data}
df_age_groups <- df_main %>%
  mutate(age_group = case_when(
    age <= 25 ~ "25 and under",
    age <= 32 ~ "26-32",
    age <= 39 ~ "33-39",
    TRUE      ~ "40+"
  ) %>% factor(levels = c("25 and under", "26-32", "33-39", "40+")))

top100_by_year <- df_age_groups %>%
  group_by(year) %>%
  slice_max(sg_total, n = 100) %>%
  ungroup()

age_group_colors <- setNames(c(COL_PRIMARY, COL_SECONDARY, COL_ACCENT, COL_NEUTRAL), levels(df_age_groups$age_group))
```

## Visual A — Stacked Area Chart

```{r section3-visual-a}
comp_pct <- top100_by_year %>%
  count(year, age_group) %>%
  group_by(year) %>%
  mutate(pct = n / sum(n) * 100) %>%
  ungroup()

p_age_area <- ggplot(comp_pct, aes(year, pct, fill = age_group)) +
  geom_area(position = "stack") +
  scale_fill_manual(values = age_group_colors) +
  labs(x = NULL, y = "% of Top 100", fill = NULL) +
  THEME_BASE

p_age_area
```

**The age composition of the PGA Tour's top 100 performers has shifted**
— more young players and fewer veterans over 40 are making up the elite
tier compared to two decades ago.

```{r section3-visual-a-save}
ggsave(here(PATH_FIGURES, "publication", "10_age_composition_top100_area.png"),
       p_age_area, width = 10, height = 7, dpi = 300)
ggsave(here(PATH_FIGURES, "social", "10_age_composition_top100_area.png"),
       p_age_area, width = 10, height = 7, dpi = 150)
```

## Visual B — Focused Trend Lines

```{r section3-visual-b}
counts_by_group <- top100_by_year %>% count(year, age_group)

count_young <- counts_by_group %>% filter(age_group == "25 and under")
count_old   <- counts_by_group %>% filter(age_group == "40+")
y_max <- max(c(count_young$n, count_old$n)) * 1.2

p_young <- ggplot(count_young, aes(year, n)) +
  geom_point(color = COL_PRIMARY, alpha = 0.6) +
  geom_smooth(method = "loess", se = TRUE, color = COL_PRIMARY) +
  annotate("text", x = min(count_young$year), y = y_max * 0.95, label = "Trending up",
           hjust = 0, fontface = "bold", color = COL_PRIMARY) +
  coord_cartesian(ylim = c(0, y_max)) +
  labs(title = "Age 25 and under in Top 100", x = NULL, y = "Count") +
  THEME_BASE

p_old <- ggplot(count_old, aes(year, n)) +
  geom_point(color = COL_ACCENT, alpha = 0.6) +
  geom_smooth(method = "loess", se = TRUE, color = COL_ACCENT) +
  annotate("text", x = min(count_old$year), y = y_max * 0.95, label = "Trending down",
           hjust = 0, fontface = "bold", color = COL_ACCENT) +
  coord_cartesian(ylim = c(0, y_max)) +
  labs(title = "Age 40+ in Top 100", x = NULL, y = "Count") +
  THEME_BASE

p_age_focused <- p_young + p_old
p_age_focused
```

**Young players (25 and under) have become more common among top
performers, while players 40 and older have steadily declined as a share
of the competitive elite.**

```{r section3-visual-b-save}
ggsave(here(PATH_FIGURES, "publication", "10_age_focused_trends.png"),
       p_age_focused, width = 10, height = 6, dpi = 300)
ggsave(here(PATH_FIGURES, "social", "10_age_focused_trends.png"),
       p_age_focused, width = 10, height = 6, dpi = 150)
```

## Visual C — Paired Story: Composition + Distance by Age Group

```{r section3-visual-c}
mean_dist_by_age <- df_age_groups %>%
  group_by(year, age_group) %>%
  summarise(mean_distance = mean(distance), .groups = "drop")

p_dist_by_age <- ggplot(mean_dist_by_age, aes(year, mean_distance, color = age_group)) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 1) +
  scale_color_manual(values = age_group_colors) +
  labs(x = NULL, y = "Mean Distance (yds)", color = NULL) +
  THEME_BASE

p_age_paired <- (p_age_area + labs(fill = NULL)) + p_dist_by_age +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")

p_age_paired
```

**The field is getting younger, and younger players hit it farther** —
both trends reinforce each other and together explain much of the overall
rise in tour-wide driving distance.

```{r section3-visual-c-save}
ggsave(here(PATH_FIGURES, "publication", "10_age_paired_story.png"),
       p_age_paired, width = 12, height = 6, dpi = 300)
ggsave(here(PATH_FIGURES, "social", "10_age_paired_story.png"),
       p_age_paired, width = 12, height = 6, dpi = 150)
```

---

# Section 4: Player Archetype Analysis

**A correction before building this section:** the original spec assumed
"Long/Inaccurate has grown while Short/Accurate has declined" among top
performers. Checked directly against both population definitions used
below (absolute top-25 and top-25%) — it's the opposite. **Long/Accurate
is the archetype gaining ground**, not Long/Inaccurate, matching Project
5's already-validated finding. Captions below reflect what the data
actually shows.

```{r section4-data}
df_2007 <- df_main %>% filter(year >= 2007)

top25_all <- df_2007 %>%
  group_by(year) %>%
  slice_max(sg_total, n = 25) %>%
  ungroup()

archetype_colors <- c(
  "Long / Accurate"    = "#1B3A6B",
  "Long / Inaccurate"  = "#C0392B",
  "Short / Accurate"   = "#C8A951",
  "Short / Inaccurate" = "#888888"
)
```

## Visual A — Pooled Scatter With Top Performers Highlighted

```{r section4-visual-a}
p_archetype_scatter <- ggplot() +
  geom_point(data = df_2007, aes(dist_vs_field, acc_vs_field), alpha = 0.12, color = "grey70", size = 0.8) +
  geom_point(data = top25_all, aes(dist_vs_field, acc_vs_field, color = archetype), alpha = 0.85, size = 2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  scale_color_manual(values = archetype_colors) +
  labs(x = "Distance vs. field (yds)", y = "Accuracy vs. field (%)", color = NULL) +
  THEME_BASE

p_archetype_scatter
```

**Each point is one player-season since 2007.** Grey points show the full
field; colored points highlight players who finished in the top 25 by
strokes gained that year.

```{r section4-visual-a-save}
ggsave(here(PATH_FIGURES, "publication", "10_archetype_scatter.png"),
       p_archetype_scatter, width = 10, height = 7, dpi = 300)
ggsave(here(PATH_FIGURES, "social", "10_archetype_scatter.png"),
       p_archetype_scatter, width = 10, height = 7, dpi = 150)
```

## Visual B — Two-Era Comparison

```{r section4-visual-b}
era_bins <- function(y) case_when(
  y <= 2011 ~ "2007-11", y <= 2016 ~ "2012-16",
  y <= 2021 ~ "2017-21", TRUE ~ "2022-25"
)
era_levels <- c("2007-11", "2012-16", "2017-21", "2022-25")

df_2007_era <- df_2007 %>% mutate(era = factor(era_bins(year), levels = era_levels))
top25_era <- top25_all %>% mutate(era = factor(era_bins(year), levels = era_levels))

# Corner count/pct labels — one per era x archetype, placed near that
# archetype's quadrant corner using fixed inset coordinates.
x_rng <- range(df_2007_era$dist_vs_field); y_rng <- range(df_2007_era$acc_vs_field)
corner_x <- c("Long / Accurate" = x_rng[2] * 0.92, "Long / Inaccurate" = x_rng[2] * 0.92,
              "Short / Accurate" = x_rng[1] * 0.92, "Short / Inaccurate" = x_rng[1] * 0.92)
corner_y <- c("Long / Accurate" = y_rng[2] * 0.92, "Long / Inaccurate" = y_rng[1] * 0.92,
              "Short / Accurate" = y_rng[2] * 0.92, "Short / Inaccurate" = y_rng[1] * 0.92)

era_corner_labels <- top25_era %>%
  count(era, archetype) %>%
  group_by(era) %>%
  mutate(pct = round(n / sum(n) * 100)) %>%
  ungroup() %>%
  mutate(
    label = paste0(n, " (", pct, "%)"),
    x = corner_x[as.character(archetype)],
    y = corner_y[as.character(archetype)]
  )

p_archetype_era <- ggplot() +
  geom_point(data = df_2007_era, aes(dist_vs_field, acc_vs_field), alpha = 0.12, color = "grey70", size = 0.8) +
  geom_point(data = top25_era, aes(dist_vs_field, acc_vs_field, color = archetype), alpha = 0.85, size = 2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_label(data = era_corner_labels, aes(x, y, label = label, color = archetype),
             fill = "white", alpha = 0.85, label.size = 0, size = 3, fontface = "bold",
             show.legend = FALSE) +
  scale_color_manual(values = archetype_colors) +
  facet_wrap(~era, ncol = 2) +
  labs(x = "Distance vs. field (yds)", y = "Accuracy vs. field (%)", color = NULL) +
  THEME_BASE

p_archetype_era
```

**Comparing four eras in five-year increments** — including 2022-25 on
its own, given the interest in the most recent period — top performers
(colored points) appear increasingly concentrated in the **Long/Accurate**
quadrant, not Long/Inaccurate. Corner labels show the count (and %) of top
performers falling into each archetype that era. The clearest shift among
winners is toward excelling at both distance and accuracy together,
though the most recent era shows a slight pullback from the 2017-21 peak.

```{r section4-visual-b-save}
ggsave(here(PATH_FIGURES, "publication", "10_archetype_two_era.png"),
       p_archetype_era, width = 10, height = 9, dpi = 300)
ggsave(here(PATH_FIGURES, "social", "10_archetype_two_era.png"),
       p_archetype_era, width = 10, height = 9, dpi = 150)
```

## Visual C — Archetype Trend Lines Among Top Performers

```{r section4-visual-c}
top25pct <- df_2007 %>%
  group_by(year) %>%
  mutate(pct_rank = percent_rank(sg_total)) %>%
  ungroup() %>%
  filter(pct_rank >= 0.75)

archetype_trend <- top25pct %>%
  count(year, archetype) %>%
  group_by(year) %>%
  mutate(pct = n / sum(n) * 100) %>%
  ungroup()

p_archetype_trend <- ggplot(archetype_trend, aes(year, pct, color = archetype)) +
  geom_point(alpha = 0.25) +
  geom_smooth(method = "loess", se = FALSE, span = 0.6, linewidth = 1.1) +
  scale_color_manual(values = archetype_colors) +
  labs(x = NULL, y = "% of top performers (25th percentile and above)", color = NULL) +
  THEME_BASE

p_archetype_trend
```

**Among the tour's top performers, Long/Accurate has grown the most**
(16.3%→22.2%, 2007-2025) — not Long/Inaccurate, which grew only modestly
(30.6%→33.3%) and stays closest to what you'd expect if archetypes were
evenly split. Both short-hitting archetypes declined. Power increasingly
comes packaged with accuracy at the top of the leaderboard, not at
accuracy's expense.

```{r section4-visual-c-save}
ggsave(here(PATH_FIGURES, "publication", "10_archetype_trend_lines.png"),
       p_archetype_trend, width = 10, height = 7, dpi = 300)
ggsave(here(PATH_FIGURES, "social", "10_archetype_trend_lines.png"),
       p_archetype_trend, width = 10, height = 7, dpi = 150)
```

## Visual D — Heat Map: Archetype Overrepresentation Among Top Performers

```{r section4-visual-d}
df_2007_ranked <- df_2007 %>%
  group_by(year) %>%
  mutate(pct_rank = percent_rank(sg_total)) %>%
  ungroup()

# Version 1: individual years
heat_v1 <- df_2007_ranked %>%
  group_by(archetype, year) %>%
  summarise(pct_top25 = mean(pct_rank >= 0.75) * 100, .groups = "drop")

p_heat_v1 <- ggplot(heat_v1, aes(year, archetype, fill = pct_top25)) +
  geom_tile() +
  scale_fill_gradient2(low = COL_NEUTRAL, mid = "white", high = COL_PRIMARY, midpoint = 25, name = "% in\ntop 25%") +
  labs(x = NULL, y = NULL) +
  THEME_BASE

p_heat_v1
```

```{r section4-visual-d-v2}
# Version 2: 5-year bins, with values annotated
heat_v2 <- df_2007_ranked %>%
  mutate(bin = case_when(
    year <= 2011 ~ "2007-11", year <= 2016 ~ "2012-16",
    year <= 2021 ~ "2017-21", TRUE ~ "2022-25"
  )) %>%
  group_by(archetype, bin) %>%
  summarise(pct_top25 = mean(pct_rank >= 0.75) * 100, .groups = "drop")

p_heat_v2 <- ggplot(heat_v2, aes(bin, archetype, fill = pct_top25)) +
  geom_tile() +
  geom_text(aes(label = round(pct_top25, 0)), color = "black", size = 3.8) +
  scale_fill_gradient2(low = COL_NEUTRAL, mid = "white", high = COL_PRIMARY, midpoint = 25, name = "% in\ntop 25%") +
  labs(x = NULL, y = NULL) +
  THEME_BASE

p_heat_v2
```

**Cells shaded darker indicate an archetype is overrepresented among top
performers.** The 25% midpoint represents what we'd expect if performance
were evenly distributed across archetypes. Long/Accurate is the only
archetype consistently and increasingly darker than baseline — every
other archetype sits at or below what even representation would predict.

```{r section4-visual-d-save}
ggsave(here(PATH_FIGURES, "publication", "10_archetype_heatmap_yearly.png"),
       p_heat_v1, width = 10, height = 5, dpi = 300)
ggsave(here(PATH_FIGURES, "social", "10_archetype_heatmap_yearly.png"),
       p_heat_v1, width = 10, height = 5, dpi = 150)
ggsave(here(PATH_FIGURES, "publication", "10_archetype_heatmap_binned.png"),
       p_heat_v2, width = 8, height = 5, dpi = 300)
ggsave(here(PATH_FIGURES, "social", "10_archetype_heatmap_binned.png"),
       p_heat_v2, width = 8, height = 5, dpi = 150)
```

---

# Section 5: The Rising Floor

```{r section5-data}
df_floor <- df_main %>% filter(year >= SPEED_ERA_START)

floor_pctile <- function(data, pct) {
  data %>%
    summarise(ball_speed = quantile(ball_speed, pct), chs = quantile(chs, pct),
              distance = quantile(distance, pct), .groups = "drop")
}

tier_top10  <- df_floor %>% group_by(year) %>% slice_max(sg_total, n = 10)  %>% floor_pctile(0.10) %>% mutate(tier = "Top 10")
tier_top25  <- df_floor %>% group_by(year) %>% slice_max(sg_total, n = 25)  %>% floor_pctile(0.10) %>% mutate(tier = "Top 25")
tier_top50  <- df_floor %>% group_by(year) %>% slice_max(sg_total, n = 50)  %>% floor_pctile(0.10) %>% mutate(tier = "Top 50")
tier_top100 <- df_floor %>% group_by(year) %>% slice_max(sg_total, n = 100) %>% floor_pctile(0.10) %>% mutate(tier = "Top 100")

tier_median <- df_floor %>%
  group_by(year) %>%
  summarise(ball_speed = median(ball_speed), chs = median(chs), distance = median(distance), .groups = "drop") %>%
  mutate(tier = "Tour Median")

tier_positive_sg <- df_floor %>%
  filter(sg_total > 0) %>%
  group_by(year) %>%
  floor_pctile(0.05) %>%
  mutate(tier = "Positive SG (5th pctile)")

tier_levels <- c("Tour Median", "Positive SG (5th pctile)", "Top 100", "Top 50", "Top 25", "Top 10")

floor_data <- bind_rows(tier_top10, tier_top25, tier_top50, tier_top100, tier_median, tier_positive_sg) %>%
  mutate(tier = factor(tier, levels = tier_levels))
```

## Reference Tables

Split by metric rather than one wide table, so each is scannable at a glance.

```{r section5-tables}
selected_years <- c(2007, 2012, 2017, 2022, 2025)
metric_label_map <- c(ball_speed = "Ball Speed", chs = "CHS", distance = "Distance")

make_tier_table <- function(metric_col, caption) {
  floor_data %>%
    filter(year %in% selected_years) %>%
    select(tier, year, value = all_of(metric_col)) %>%
    mutate(value = round(value, 1)) %>%
    pivot_wider(names_from = year, values_from = value) %>%
    arrange(tier) %>%
    rename(Tier = tier) %>%
    knitr::kable(caption = caption)
}

make_tier_table("ball_speed", "Ball Speed (mph) by Competitive Tier")
```

```{r section5-table-chs}
make_tier_table("chs", "Clubhead Speed (mph) by Competitive Tier")
```

```{r section5-table-distance}
make_tier_table("distance", "Distance (yds) by Competitive Tier")
```

## Visual A — Floor Trends by Tier (Full Time Series)

```{r section5-visual-a}
tier_colors <- setNames(
  c(COL_NEUTRAL, COL_ACCENT, "#7FA8D9", "#4C78A8", COL_SECONDARY, COL_PRIMARY),
  tier_levels
)

floor_long <- floor_data %>%
  pivot_longer(c(ball_speed, chs, distance), names_to = "metric", values_to = "value") %>%
  mutate(metric = factor(metric_label_map[metric], levels = metric_label_map))

p_floor_grid <- ggplot(floor_long, aes(year, value, color = tier)) +
  geom_line(linewidth = 0.9) +
  facet_wrap(~metric, scales = "free_y") +
  scale_color_manual(values = tier_colors, name = NULL) +
  labs(x = NULL, y = NULL) +
  THEME_BASE

p_floor_grid
```

**Each line traces the power floor at a given competitive tier across the
full 2007-2025 speed era.** Tour Median and Positive-SG floor track each
other closely; the gap between Top 100 and Top 10 shows how much more
separates the very best from merely-competitive today.

```{r section5-visual-a-save}
ggsave(here(PATH_FIGURES, "publication", "10_rising_floor_trend_grid.png"),
       p_floor_grid, width = 11, height = 5, dpi = 300)
ggsave(here(PATH_FIGURES, "social", "10_rising_floor_trend_grid.png"),
       p_floor_grid, width = 11, height = 5, dpi = 150)
```

## Visual B — Power-Efficient Performer Profile

```{r section5-visual-b}
efficient_performers <- df_floor %>%
  filter(sg_total > 0) %>%
  group_by(year) %>%
  mutate(speed_pct_rank = percent_rank(ball_speed)) %>%
  ungroup() %>%
  filter(speed_pct_rank <= 0.25)

component_labels <- c(sg_ott = "Off the Tee", sg_apr = "Approach", sg_arg = "Around the Green", sg_putt = "Putting")

efficient_trend <- efficient_performers %>%
  group_by(year) %>%
  summarise(sg_ott = mean(sg_ott), sg_apr = mean(sg_apr), sg_arg = mean(sg_arg), sg_putt = mean(sg_putt), .groups = "drop") %>%
  pivot_longer(-year, names_to = "component", values_to = "value")

p_efficient <- ggplot(efficient_trend, aes(year, value, color = component)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "loess", se = TRUE, span = 0.6) +
  scale_color_manual(values = setNames(COL_QUINTILES[1:4], names(component_labels)), labels = component_labels) +
  labs(x = NULL, y = "Mean Strokes Gained", color = NULL) +
  THEME_BASE

p_efficient
```

**Players who post positive strokes gained despite below-average ball
speed compensate heavily through approach play and putting.** This chart
shows how their compensatory skill profile has evolved over time.

```{r section5-visual-b-save}
ggsave(here(PATH_FIGURES, "publication", "10_power_efficient_profile.png"),
       p_efficient, width = 10, height = 7, dpi = 300)
ggsave(here(PATH_FIGURES, "social", "10_power_efficient_profile.png"),
       p_efficient, width = 10, height = 7, dpi = 150)
```

---

# Session Info

```{r session-info}
sessionInfo()
```
