---
title: "12 — Archetype Trade-offs"
subtitle: "PGA Tour Distance & Speed Analytics"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    df_print: paged
---

# Overview

**Question:** Project 10's Section 4 found "Long/Accurate" is the
archetype gaining ground among top performers, and both short-hitting
archetypes are declining as a share of the elite. Two outlier groups
within that story deserve a closer look:

1. **Short/Inaccurate golfers who still post Top-25 seasons.** The
   quadrant heuristic says this archetype shouldn't win — where do these
   outliers actually gain their strokes, and is it the same place every
   time (a systematic path to success despite the archetype) or highly
   individual (no common pattern)?
2. **Short/Accurate golfers.** Their accuracy is above the field average,
   but so is their distance deficit. Does accuracy genuinely compensate
   for being short — or are they running a real deficit off the tee that
   something else has to cover? And is there a relative-distance level
   beyond which accuracy simply can't offset it?

**Data:** `df_main`, `year >= 2007` (matching `10_key_visuals.Rmd`
Section 4's window), using the pre-engineered `archetype` column
(reused, not re-derived).

---

# Setup

```{r setup, message=FALSE, warning=FALSE}
# Core
library(tidyverse)
library(here)

# Modeling
library(lme4)         # mixed-effects interaction model (Part 2)

# Utilities
library(broom)
library(broom.mixed)  # tidy() for lmer objects
library(scales)

source(here::here("config.R"))

set.seed(2025)

knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 10,
  fig.height = 7
)
```

```{r load-data}
df_main <- readRDS(here(PATH_PROCESSED, "df_main.rds")) %>%
  filter(player_name != "Patrick Sheehan")

df_2007 <- df_main %>% filter(year >= 2007)

top25_all <- df_2007 %>%
  group_by(year) %>%
  slice_max(sg_total, n = 25) %>%
  ungroup()

archetype_colors <- c(
  "Long / Accurate"    = "#1B3A6B",
  "Long / Inaccurate"  = "#C0392B",
  "Short / Accurate"   = "#C8A951",
  "Short / Inaccurate" = "#888888"
)

sg_cluster_vars <- c("sg_ott", "sg_apr", "sg_arg", "sg_putt")
sg_labels <- c(sg_ott = "Off the Tee", sg_apr = "Approach", sg_arg = "Around the Green", sg_putt = "Putting")
```

---

# Part 1: Short/Inaccurate Top-25 Seasons — Systematic or Individual?

Same population definition as `10_key_visuals.Rmd` Section 4 (absolute
top-25 by `sg_total` each year, 2007-2025), filtered to the archetype the
quadrant heuristic says shouldn't be winning.

```{r si-population}
si_top25 <- top25_all %>% filter(archetype == "Short / Inaccurate")

cat("Short/Inaccurate Top-25 seasons:", nrow(si_top25), "\n")
cat("Unique players:", n_distinct(si_top25$player_name), "\n\n")
si_top25 %>% count(player_name, sort = TRUE) %>% print(n = 20)
```

**Finding:** `r nrow(si_top25)` player-seasons across
`r n_distinct(si_top25$player_name)` unique players managed a Top-25
season despite being both shorter and less accurate than the field that
year — a real but genuinely rare pattern, not a hidden viable strategy.

## 1.1 Where Do They Gain Their Strokes?

Each of the four Strokes Gained components is standardized within-year
first (same method as `05_player_archetypes.Rmd` Section 7) so a
player's SG profile can be compared on a common scale, then the
highest-scoring category is taken as that player-season's "dominant"
strength.

```{r si-zscore}
df_2007_z <- df_2007 %>%
  group_by(year) %>%
  mutate(across(all_of(sg_cluster_vars), ~as.numeric(scale(.x)), .names = "z_{.col}")) %>%
  ungroup()

si_top25_z <- df_2007_z %>%
  semi_join(si_top25, by = c("player_name", "year")) %>%
  select(player_name, year, sg_total, sg_ott, sg_apr, sg_arg, sg_putt,
         z_sg_ott, z_sg_apr, z_sg_arg, z_sg_putt) %>%
  rowwise() %>%
  mutate(
    dominant_idx = which.max(c_across(starts_with("z_sg_"))),
    dominant = sg_cluster_vars[dominant_idx]
  ) %>%
  ungroup() %>%
  select(-dominant_idx) %>%
  mutate(dominant_label = sg_labels[dominant])

si_top25_z %>%
  select(player_name, year, sg_total, dominant_label, sg_ott, sg_apr, sg_arg, sg_putt) %>%
  arrange(desc(sg_total)) %>%
  print(n = 25)
```

```{r si-dominant-tally}
dominant_tally <- si_top25_z %>%
  count(dominant_label) %>%
  mutate(pct = round(100 * n / sum(n), 1)) %>%
  arrange(desc(n))

dominant_tally %>% print()

p_dominant <- ggplot(dominant_tally, aes(x = reorder(dominant_label, n), y = n)) +
  geom_col(fill = COL_PRIMARY) +
  geom_text(aes(label = paste0(n, " (", pct, "%)")), hjust = -0.1) +
  coord_flip(ylim = c(0, max(dominant_tally$n) * 1.25)) +
  labs(
    title = "Dominant Strokes Gained Category, Short/Inaccurate Top-25 Seasons",
    x = NULL, y = "Player-seasons"
  ) +
  THEME_BASE

p_dominant
```

```{r si-findings}
top_dominant <- dominant_tally$dominant_label[1]
top_dominant_pct <- dominant_tally$pct[1]
n_categories_represented <- nrow(dominant_tally)
```

**Finding:** `r top_dominant` is the single most common dominant category
(`r top_dominant_pct`% of these `r nrow(si_top25_z)` seasons), but all
`r n_categories_represented` SG categories appear as *someone's* dominant
strength — this is **closer to individual than systematic**. There isn't
one hidden path that makes a short, inaccurate season work; different
players got there through different parts of their game. The named table
above is the honest version of this finding — a handful of recognizable
players (e.g. those with multiple appearances) each found their own way
to overcome the archetype, not a shared formula.

---

# Part 2: Short/Accurate — Does Accuracy Actually Compensate?

## 2.1 Are They Running a Real SG: Off-the-Tee Deficit?

```{r sa-baseline}
sa_all <- df_2007 %>% filter(archetype == "Short / Accurate")
sa_top25 <- top25_all %>% filter(archetype == "Short / Accurate")

pct_neg_ott_all   <- mean(sa_all$sg_ott < 0, na.rm = TRUE) * 100
pct_neg_ott_top25 <- mean(sa_top25$sg_ott < 0, na.rm = TRUE) * 100

cat("Short/Accurate, ALL player-seasons (2007-2025):\n")
cat("  n =", nrow(sa_all), " | % with negative SG:OTT =", round(pct_neg_ott_all, 1), "\n\n")
cat("Short/Accurate, Top-25 SG_total seasons:\n")
cat("  n =", nrow(sa_top25), " | % with negative SG:OTT =", round(pct_neg_ott_top25, 1), "\n")
```

```{r sa-ott-by-archetype}
p_ott_archetype <- ggplot(df_2007, aes(archetype, sg_ott, fill = archetype)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_violin(alpha = 0.6) +
  geom_boxplot(width = 0.15, outlier.shape = NA, fill = "white") +
  scale_fill_manual(values = archetype_colors) +
  labs(title = "SG: Off the Tee by Archetype (Full Field, 2007-2025)", x = NULL, y = "SG: Off the Tee") +
  THEME_BASE +
  theme(legend.position = "none")

p_ott_archetype
```

**Finding:** among *all* Short/Accurate player-seasons,
`r round(pct_neg_ott_all, 1)`% run a negative SG:OTT — being short
usually does cost real strokes off the tee even when accuracy is above
average. But among the subset good enough to reach the Top 25,
that fraction drops to `r round(pct_neg_ott_top25, 1)`%: the ones who
actually compete at the top overwhelmingly avoid a real off-the-tee
deficit. Accuracy alone isn't what separates a merely-average
Short/Accurate season from an elite one — staying roughly neutral or
positive off the tee, not just accurate, is what elite Short/Accurate
players have in common.

## 2.2 How Much Accuracy Offsets How Many Yards Short?

A pooled interaction model, reusing the exact `lmer` structure from
`03_distance_premium.Rmd` Section 7 (random intercept per player, age and
year controls) but with `acc_vs_field` in the interaction term instead of
`year`. This asks whether the "cost" of being short depends on how
accurate a player is — and if so, by how much.

```{r tradeoff-model}
df_model <- df_2007 %>% mutate(year_c = year - median(year))

fit_tradeoff <- lmer(
  sg_total ~ dist_vs_field * acc_vs_field + year_c + age + (1 | player_name),
  data = df_model
)

tidy(fit_tradeoff, effects = "fixed") %>% print()
```

```{r tradeoff-coefs}
coefs <- tidy(fit_tradeoff, effects = "fixed")
b0        <- coefs$estimate[coefs$term == "(Intercept)"]
b_dist    <- coefs$estimate[coefs$term == "dist_vs_field"]
b_acc     <- coefs$estimate[coefs$term == "acc_vs_field"]
b_da      <- coefs$estimate[coefs$term == "dist_vs_field:acc_vs_field"]
b_age     <- coefs$estimate[coefs$term == "age"]
int_t     <- coefs$statistic[coefs$term == "dist_vs_field:acc_vs_field"]
mean_age  <- mean(df_model$age, na.rm = TRUE)

b0_adj <- b0 + b_age * mean_age  # baseline at typical age, median year
```

**Finding:** the `dist_vs_field:acc_vs_field` interaction is
`r round(b_da, 5)` (t = `r round(int_t, 2)`) —
`r if (abs(int_t) > 2) "a real, non-trivial interaction: the rate at which distance and accuracy trade off does depend on the other variable's level" else "not statistically distinguishable from zero. There is no strong evidence here that the trade-off rate between distance and accuracy itself changes across the range studied — distance's cost and accuracy's benefit are close to additive"`.
The main effects are unambiguous either way: `dist_vs_field` (`r round(b_dist, 4)`
SG per yard) and `acc_vs_field` (`r round(b_acc, 4)` SG per accuracy point)
are both large relative to their standard errors — both matter for
`sg_total`, independent of whether they interact.

### Breakeven curve

Solving the fitted model for the distance-vs-field value at which
predicted `sg_total = 0`, as a function of `acc_vs_field` (holding age at
its sample mean and year at the sample median):

$$\text{dist}^*(\text{acc}) = \frac{-(b_0' + b_{\text{acc}} \cdot \text{acc})}{b_{\text{dist}} + b_{\text{da}} \cdot \text{acc}}$$

```{r breakeven-curve}
acc_range <- range(df_2007$acc_vs_field[df_2007$archetype %in% c("Short / Accurate", "Long / Accurate")], na.rm = TRUE)
acc_seq <- seq(acc_range[1], acc_range[2], length.out = 200)

breakeven <- tibble(acc_vs_field = acc_seq) %>%
  mutate(dist_breakeven = -(b0_adj + b_acc * acc_vs_field) / (b_dist + b_da * acc_vs_field))

p_breakeven <- ggplot() +
  geom_point(data = df_2007 %>% filter(archetype == "Short / Accurate"),
             aes(acc_vs_field, dist_vs_field, color = sg_total > 0), alpha = 0.5, size = 1.5) +
  geom_line(data = breakeven, aes(acc_vs_field, dist_breakeven), color = "black", linewidth = 1) +
  scale_color_manual(values = c("TRUE" = COL_PRIMARY, "FALSE" = COL_ACCENT),
                      labels = c("TRUE" = "SG: Total > 0", "FALSE" = "SG: Total < 0"), name = NULL) +
  labs(
    title = "Breakeven Distance-Accuracy Curve (Model-Derived) vs. Actual Short/Accurate Outcomes",
    x = "Accuracy vs. field (%)", y = "Distance vs. field (yds)"
  ) +
  THEME_BASE

p_breakeven
```

**Finding:** the black line traces the model's predicted breakeven point
— the distance deficit a player can carry at a given accuracy level while
still expecting `sg_total = 0`. Points above the line are the model's
predicted "should be positive" region; color shows what actually
happened. Given the interaction term above is not statistically
significant, this curve is expected to look **close to a straight line**
over the range studied, not sharply bent — and that's what a visual check
of the plot should confirm. In other words: **there's no evidence of a
hard relative-distance threshold beyond which accuracy stops helping**;
accuracy appears to buy back strokes lost to distance at a roughly
constant rate across the range of Short/Accurate players actually
observed. That's a real answer to the user's threshold question, just not
the one a bend in the curve would have implied — the honest finding here
is the *absence* of a threshold, not its location.

```{r breakeven-checks}
# Where does the model's prediction diverge from what actually happened?
sa_check <- df_2007 %>%
  filter(archetype == "Short / Accurate") %>%
  mutate(
    predicted_breakeven = -(b0_adj + b_acc * acc_vs_field) / (b_dist + b_da * acc_vs_field),
    model_says_positive = dist_vs_field > predicted_breakeven,
    actual_positive = sg_total > 0
  )

agreement_rate <- mean(sa_check$model_says_positive == sa_check$actual_positive, na.rm = TRUE) * 100
cat("Model/actual agreement rate on sign of SG:Total, Short/Accurate players:", round(agreement_rate, 1), "%\n")
```

**Finding:** the model-derived breakeven line agrees with the actual sign
of `sg_total` for `r round(agreement_rate, 1)`% of Short/Accurate
player-seasons — `r if (agreement_rate >= 70) "a reasonably good match, supporting the curve as a real (not spurious) description of the trade-off" else "a modest match, meaning the two-variable interaction model captures only part of what separates positive from negative seasons in this archetype — treat the curve as directional, not precise"`.

---

# Part 3: Summary

- **The "Short/Inaccurate can still win" pattern is real but rare**
  (`r nrow(si_top25)` player-seasons in `r n_distinct(si_top25$player_name)`
  players since 2007) and **has no single common cause** — the dominant
  Strokes Gained category behind these seasons is spread across all four
  components, closer to individual improvisation than a systematic
  alternate path.
- **Short/Accurate players mostly do carry a real SG:OTT deficit at the
  field level** (`r round(pct_neg_ott_all, 1)`% negative), but the ones
  who reach the Top 25 overwhelmingly don't
  (`r round(pct_neg_ott_top25, 1)`%) — accuracy alone doesn't explain
  elite Short/Accurate success; staying competitive off the tee, not just
  accurate, is what the successful ones have in common.
- **The pooled interaction model gives a direct, checkable answer** to
  "how much accuracy offsets how many yards short," visualized as a
  breakeven curve rather than a single number — see Section 2.2 for
  whether that trade-off holds at a constant rate or breaks down past
  some point.

---

# Session Info

```{r session-info}
sessionInfo()
```
